Noncanonical usage of stop codons in ciliates expands proteins with structurally flexible Q-rich motifs

Version 1

Abstract

Abstract

Serine(S)/threonine(T)-glutamine(Q) cluster domains (SCDs), polyglutamine (polyQ) tracts and polyglutamine/asparagine (polyQ/N) tracts are Q-rich motifs found in many proteins. SCDs often are intrinsically disordered regions that mediate protein phosphorylation and protein-protein interactions. PolyQ and polyQ/N tracts are structurally flexible sequences that trigger protein aggregation. We show that four SCDs and three prion-causing Q/N-rich motifs of yeast proteins possess autonomous protein expression-enhancing activities. Comparative Gene Ontology (GO) analyses of the near-complete proteomes of 27 representative model eukaryotes reveal that Q-rich motifs prevail in proteins involved in specialized biological processes, including Saccharomyces cerevisiae RNA-mediated transposition, Candida albicans filamentous growth, ciliate peptidyl-glutamic acid modification, Tetrahymena thermophila xylan catabolism and meiosis, Dictyostelium discoideum development and sexual cycles, Plasmodium falciparum infection, and the Drosophila melanogaster nervous system. We also show that Q-rich motifs are expanded massively in ten ciliates with reassigned TAAQ and TAGQ codons. Our results provide new insights to explain why many ciliates reassign their nuclear stop codons into glutamine (Q). The consequence of this preponderance of Q is massive expansion of proteins harboring three structurally flexible or even intrinsically disordered Q-rich motifs. Since these Q-rich motifs can endow proteins with structural and functional plasticity, we suggest that they represent useful toolkits for evolutionary novelty.

Introduction

Ciliated protists (ciliates) are a large and diverse group of microbial eukaryotes that are important to a wide range of biological studies and applications. Relative to most other eukaryotes, the nuclear genomes of many ciliates possess three interesting features. First, ciliates often have two types of nuclei. Their diploid micronucleus (MIC) carries the cell germline, the genetic material of which is inherited via sexual reproduction and meiosis. The polyploid macronucleus (MAC) or vegetative nucleus provides nuclear RNA for vegetative growth. MAC is generated from the MIC by massive amplification, editing and rearrangement of the genome (see reviews in (1, 2)). Second, ciliate meiosis is remarkable relative to that of other studied sexual eukaryotes. A hallmark of meiosis is programmed initiation of DNA double-strand breaks (DSBs) by Spo11, a meiosis-specific topoisomerase VII subunit A endonuclease (3, 4). Meiotic DSBs are then repaired in a homology-dependent manner via searching, pairing (or synapsis) and exchange of homologous parental chromosomes, leading to the formation of interhomolog crossovers (COs). Ultimately, the resulting recombinant chromosomes equally segregate into haploid gametes. To facilitate pairing and recombination between homologous chromosomes, telomeres attach to the nuclear envelope (NE) at the beginning of meiosis and then cluster to one side of the NE to form a transient structure called the “bouquet”. In many sexual eukaryotes, homology pairing usually culminates in the synaptonemal complex (SC), a zipper-like proteinaceous structure that appears to regulate CO formation. The SC behaves like a liquid crystal, tightly connecting paired homologous chromosomes from telomere to telomere and also incorporating lateral component loops in interstitial chromosome segments (5, 6). Most studied ciliates lack SC or only have degenerate SCs (7). In Tetrahymena thermophila, the most intensively studied ciliate, meiotic MICs undergo extreme elongation (by ∼50-fold) and form proteinaceous condensates called “crescents”. Within these elongated crescents, telomeres and centromeres of all meiotic chromosomes are rearranged at opposing ends in a stretched bouquet-like manner. Meiotic pairing and recombination take place within the crescents (see review in (8)). It has been reported that ATR1 (Ataxia Telangiectasia Mutated 1), an evolutionarily conserved DNA damage senor protein kinase, senses Spo11-induced DSBs and triggers the elongation of MICs (9). Meiosis-specific CYC2 and CYC17 cyclins, as well as cyclin-dependent kinase CDK3, are required to initiate meiosis and for crescent assembly (10-12). CYC2/CDK2 promotes bouquet formation in MIC via controlling microtubule-directed elongation (12) and also controls gene expression of proteins involved in DSB formation (SPO11), DNA repair (COM1, EXO1, DMC1), and CO formation (HOP2, MND1, MSH4, MSH5, ZPH3, BIM1, and BIM2) (13). The DPL2/E2fl1 complex, a meiosis-specific transcription factor, promotes transcriptional induction of DNA repair proteins and chromosome-associated structural proteins, including MRE11, COM1, EXO1, RAD50, RAD51, SMC1, SMC2, SMC3, SMC4, REC8, ESP1, and CNA1, among others (13). Nevertheless, the molecular mechanisms underlying crescent assembly and disassembly remain poorly understood. Third, many ciliates reassign their standard stop codons into amino acids (Table 1). For example, several ciliates possess two noncanonical nuclear genetic codes (UAAQ and UAGQ), in which the UAA and UAG stop codons have been reassigned to glutamine (Q) so that UGA is the sole functional stop codon in many ciliates, including Tetrahymena thermophila, Paramecium tetraurelia, Paramecium bursaria, Oxytricha trifallax, Stylonychia lemnae, Pseudocohnilembus persalinus, Aristerostoma sp., Favella ehrenbergii, Pseudokeronopsis spp., Strombidium inclinatum, and Uronema spp. Both the UAA and UAG stop codons are reassigned to tyrosine (Y) in Favella ehrenbergii, whereas the UGA stop codon is translated to tryptophan (W) or cysteine (C) in Blepharisma japonicum and Euplotes focardii, respectively. In contrast, Stentor coeruleus, Climacostomum virens, Litonotus pictus and Protocruzia adherens utilize the universal set of genetic codons. Notably, Condylostoma magnum and Parduzcia sp. have no dedicated genetic codes. Their UAA, UAG and UGA codons can be stop codons or translated to Q, C and W, respectively. Translation termination at the mRNA 3′ end occurs in a context-dependent manner to distinguish stop from sense (14-26).

Although it has been reported previously that Q is used more frequently in Tetrahymena thermophila and Paramecium tetraurelia than in other species (19, 20), many important questions regarding stop codon reassignment in ciliates remain unresolved. For instance, fundamentally, it is unclear if Q, Y, W and C are used more frequently in other ciliates in which stop codons are reassigned. Moreover, whether there are common or specific structural motif(s) in proteins arising from stop codon reassignment is not clear. Furthermore, what are the structural and functional impacts of such genome-wide alterations? Since high-quality transcriptomic and gene annotation data are now available for many ciliate and non-ciliate eukaryotes and genetic codons have also been properly defined for a wide variety of them (Table 1), we were confident we could provide important insights into these interesting topics.

In this report, first we present an unexpected result that the Q-rich motifs of several Saccharomyces cerevisiae proteins possess autonomous protein expression-enhancing (PEE) activities (27), including four serine(S)/threonine(T)-Q cluster domains (SCDs) and three prion-forming domains (PFDs) with polyglutamine (polyQ) or polyglutamine/asparagine (polyQ/N) tracts. Notably, all seven of these Q- or Q/N-rich motifs display high Q, S, T, and/or N contents. A common feature of intrinsically disordered regions (IDRs) is high content of Q, N, S, T, proline (P), glycine (G) and charged amino acids. Rather than adopting stable secondary and/or tertiary structures, IDRs are structurally flexible polypeptides in their native states (28-30). Next, we performed Gene Ontology (GO) enrichment analyses on all proteins having Q-rich motifs (i.e., SCDs, polyQ and polyQ/N), as well as those with the homorepeat (polyX) motifs of other amino acid residues, in 20 non-ciliate and 17 ciliate species. Our results support that Q-rich motifs are not only massively expanded in the 10 ciliate species we examined with reassigned TAAQ and TAGQ codons, but they also prevail in proteins involved in several species-specific or even phylum-specific biological processes.

Results

SCDs provide versatile functionalities in proteins

We reported previously that the NH2-terminal domain (NTD; residues 1–66) of budding yeast Saccharomyces cerevisiae Rad51 protein (Rad51-NTD) autonomously promotes high-level production of native Rad51 and its COOH-terminal fusion protein LacZ (β-galactosidase) in vivo (27). In brief, we expressed Rad51-LacZ-NVH fusion proteins using a CEN-ARS plasmid (low-copy number) under the control of the native RAD51 gene promoter (PRAD51) (Table 1). The NVH tag contains an SV40 nuclear localization signal (NLS) peptide preceding a V5 epitope tag and a hexahistidine (His6) affinity tag (27). We confirmed that N-terminal addition of Rad51-NTD to LacZ-NVH increased both steady-state levels of LacZ-NVH fusion proteins (Figure 1A) and β-galactosidase activities in vivo (Figure 1B).

The Q-rich domains of seven different yeast proteins possess autonomous expression-enhancing (PEE) activities.

(A-B) N-terminal fusion of Rad51-NTD/SCD, Rad53-SCD1, Hop1-SCD, Sml1-NTD, Sup35-PND, Ure2-UPD and New1-NPD promotes high-level expression of LacZ-NVH, respectively. The NVH tag contains an SV40 nuclear localization signal (NLS) peptide preceding a V5 epitope tag and a hexahistidine (His6) affinity tag (27). Western blots for visualization of LacZ-NVH fusion proteins (A) and quantitative β-galactosidase assays (B) were carried out as described previously (27). Error bars indicate standard deviation between experiments (n ≥ 3). Asterisks indicate significant differences relative to wild type (WT) in A or lacking an NTD in B, with P values calculated using a two-tailed t-test (***, P value <0.001; **, P value <0.01). (C-D) The PEE activities of S/T/Q/N-rich domains are independent of the quaternary structures of target proteins. (C) Rad53-SCD1 can be used as an N-terminal fusion tag to enhance production of four different target proteins: LacZ-NVH, GST-NVH, GSTnd-NVH and GFP-NVH. (D) Visualization of native Rad51 (NTD-Rad51-ΔN), Rad51-ΔN, and the Rad51-ΔN fusion proteins by immunoblotting. Hsp104 was used as a loading control. Size in kilodaltons of standard protein markers is labeled to the left of the blots. The black arrowhead indicates the protein band of Rad51-ΔN. (E) MMS sensitivity. Spot assay showing five-fold serial dilutions of indicated strains grown on YPD plates with or without MMS at the indicated concentrations (w/v).

Rad51-NTD is a SCD because it has three SQ motifs (S2Q, S12Q and S30Q) (27). The S/T-Q motifs, comprising S or T followed by Q, are the target sites of DNA damage senor protein kinases, i.e., ATM, ATR (ATM and RAD3-related) and DNA-dependent protein kinase (DNA-PK) (31, 32). Mec1 (Mitotic Entry Checkpoint 1) and Tel1 (TELomere maintenance 1) are the budding yeast homologs of mammalian ATR and ATM, respectively. Budding yeast lacks a DNA-PK homolog (33-35). We showed that the three SQ motifs of Rad51-NTD are phosphorylated by Mec1 and Tel1 during vegetative growth and meiosis. Mec1/Tel1-dependent NTD phosphorylation antagonizes Rad51 degradation via the proteasomal pathway, increasing the half-life of Rad51 from ∼30 min to ≥180 min (27). This outcome is consistent with the notion that Mec1 exhibits an essential function in regulating protein homeostasis (proteostasis) in S. cerevisiae (36, 37).

A unifying definition of an SCD is possession of three or more S/T-Q sites within a stretch of ≤50 or ≤100 amino acids in S. cerevisiae and human, respectively (31, 32). One of the best-understood mechanisms of SCD phosphorylation involves their association with binding partners hosting a forkhead-associated (FHA) domain. For example, Mec1 and Tel1 phosphorylate the SCD1 of Rad53 (residues 1-29) and the SCD of Hop1 (residues 258-324), respectively, specifically recruiting and activating their downstream DNA damage checkpoint kinases Dun1 and Mek1 during vegetative growth and meiosis (38-40). Sml1, an inhibitor of ribonucleotide reductase, is involved in regulating Rad53 and the production of deoxynucleoside triphosphate (dNTP) during DNA replication and the DNA damage response (DDR) (41). The Sml1 protein in the SK1 strain harbors three S/T-Q motifs (S4Q, S14Q and T47Q), whereas that in the S288c strain only has one SQ motif (S4Q, C14Q and T47M). Here, we further show that yeast Rad53-SCD1, Hop1-SCD, Sml1-NTD1– 50 (residues 1–50) and Sml1-NTD1–27 (residues 1–27) also exhibit PEE activities (Figure 1A and 1B, Table 1).

The Q-rich motifs of three yeast prion-causing proteins also exhibit PEE activities

Since Sml1-NTD1–27 in the SK1 strain only harbors two S/T-Q motifs (S4Q and S14Q), the number of S/T-Q motifs alone could not account for PEE activity. Notably, Rad51-NTD, Rad53-SCD1, Hop1-SCD, Sml1-NTD1-27 and Sml1-NTD1-50 all represent Q- or Q/N-rich motifs. Rad51-NTD contains 9 serines (S), 2 threonines (T), 9 glutamines (Q) and 4 asparagines (N). Rad53-SCD1 has 2S, 4T, 7Q and 1N. Hop1-SCD has 6S, 6T, 8Q and 9N. Sml1-NTD1–27 and Sml1-NTD1–50 in SK1 possess 3S and 5S, 2T and 3T, 6Q and 7Q, as well as 2N and 3N, respectively.

We investigated if other Q- or Q/N-rich motifs in yeast can also promote protein expression in vivo. PolyQ and polyQ/N tracts are the most common homorepeats acting as structurally flexible motifs for protein aggregation or protein–protein interactions in many eukaryotes (42, 43). PolyN is not as structurally flexible as polyQ due to a stronger propensity for β-turn formation in polyN than in polyQ (44). In so-called polyQ-associated diseases, long Q-, Q/N- or even N-rich motifs result in an excess of interactions, resulting in malfunctional or pathogenic protein aggregates (45). Many prion-causing proteins contain Q/N-rich prion forming domains (PFDs). In S. cerevisiae, the best-characterized prion-causing proteins are Sup35 (or translation terminator eRF35), New1 ([NU+] prion formation protein 1), Ure2 (uridosuccinate transport 2), Rnq1 (rich in N and Q 1) and Swi1 (switching deficient 1) (46, 47). We found that the Q/N-rich NTDs of Sup35, Ure2 and New1 also display PEE activities, i.e., the prion nucleation domain (PND; residues 1-39) of Sup35, the Ure2 prion domain (UPD) (residues 1-91) and the New1 prion domain (NPD; residue 1-146) (Figure 1A and 1B, Table 1). Sup35-PND containing 3S, 12Q, 18N and an S17Q motif exerts critical functions in promoting [PSI+] prion nucleation (48). The UPD of the Ure2 nitrogen catabolite repression transcriptional regulator is the basis of the prion [URE3+] (49, 50). The UPD is critical for Ure2’s function in vivo because its removal in the corresponding Ure2-ΔUPD mutants elicits reduced protein stability and steady-state protein levels (but not transcript levels) (51). Ure2-UPD contains 10S, 5T, 10Q and 33N, adopting a completely disordered structure (52). New1 is a non-essential ATP-binding cassette type F protein that fine-tunes the efficiency of translation termination or ribosome recycling (53). The NPD of New1 supports [NU+] and is susceptible to [PSI+] prion induction (54, 55). New1-NPD contains 19S, 8T, 14Q, 28N and an S145Q motif. Here, we applied the LacZ-NVH fusion protein approach to show that N-terminal fusion of Sup35-PND, Ure2-UPD or New1-NPD to LacZ-NVH all increased steady-state protein levels (Figure 1A) and β-galactosidase activities in vivo (Figure 1B).

The PEE function is not affected by the quaternary structures of target proteins

We found that N-terminal fusion of Rad53-SCD1 to four different NVH-tagged target proteins (Figure 1C) or Rad51-ΔN (Figure 1D) all resulted in higher protein production in vivo. LacZ is a tetrameric protein, glutathione S-transferase (GST) is dimeric, and non-dimerizing GST (GSTnd) and GFP are monomeric proteins. As reported recently (27), removal of the NTD from Rad51 reduced by ∼97% the protein levels of corresponding Rad51-ΔN proteins relative to wild type (WT) (Figure 1D), leading to lower resistance to the DNA damage agent methyl methanesulfonate (MMS) (Figure 1E). Interestingly, the autonomous PEE function of Rad51-NTD could be fully rescued in Rad51-ΔN by N-terminal fusion of Rad53-SCD1, Rad53-SCD1-5STA (all five S/T-Q motifs changed to AQs) or Sup35-PND, respectively. Compared to wild-type yeast cells, the three corresponding yeast mutants (rad51-SCD1-rad51-ΔN, rad51-SCD1-5STA-rad51-ΔN and sup35-PND-rad51-ΔN) (Table 2) not only produced similar steady-state levels of Rad51- ΔN fusion proteins (Figure 1D), but they also exhibited high MMS resistance (Figure 1E). During homology directed repair of DNA double strand breaks (DSBs), Rad51 polymerizes into helical filaments on DSB-associated single-stranded DNA (ssDNA) and then promotes homologous search and strand exchange of the ssDNA-protein filament with a second double-stranded DNA (dsDNA). We inferred that the catalytic activity of Rad51- ΔN in HDRR is likely similar to that of wild-type Rad51 because the weak MMS-resistant phenotype of rad51-ΔN is mainly due to very low steady-state levels of Rad51-ΔN (Figure 1D). Accordingly, we concluded that the quaternary structures of the target proteins (i.e., GFP, GSTnd, GST, LacZ and Rad51-ΔN) are irrelevant to the autonomous PEE activity.

The autonomous PEE function is not likely controlled by plasmid copy number or its transcription

We also show that the PEE function is unlikely to operate at the transcriptional level, as revealed by genomic and reverse-transcription quantitative polymerase chain reactions (i.e., g-qPCR and RT-qPCR, respectively) (Figure 2 and Table 3). We found that addition of WT and mutant Rad51-NTD to LacZ-NVH not only did not affect the average copy number of the corresponding CEN-ARS plasmids in exponentially growing S. cerevisiae cells (Figure 2A), but also even reduced the steady-state transcript levels of the corresponding LacZ-NVH fusion protein genes (Figure 2B). Therefore, addition of Rad51-NTD to LacZ-NVH did not result in a significant increase in transcription.

The autonomous protein-expression-enhancing function of Rad51-NTD is unlikely to be controlled during transcription or simply arise from plasmid copy number differences.

The effects of WT and mutant Rad51-NTD on β-galactosidase activities (A), plasmid DNA copy numbers (B), relative steady-state levels of LacZ-NVH mRNA normalized to ACT1 (actin) mRNA (C), and relative ratios of LacZ-NVH mRNA versus plasmid DNA copy number (D). The wild-type yeast cells were transformed with indicated CEN-ARS plasmids, respectively, to express WT and mutant Rad51-NTD-LacZ-NVH fusion proteins or LacZ-NVH alone under the control of the native RAD51 gene promoter (PRAD51). The relative quantification (RQ = 2-ΔΔϹT) values were determined to reveal the plasmid DNA copy number and steady-state levels of LacZ-NVH mRNA by g-qPCR and RT-qPCR, respectively. LacZ and ACT1 were selected as target and reference protein-encoding genes, respectively, in both g-qPCR and RT-qPCR. The data shown represent mean ± SD from three independent biological data-points.

The protein quality control system moderately regulates autonomous PEE activities

The protein quality control system is a mechanism by which cells monitor proteins to ensure that they are appropriately folded (56). In the current study, we compared the protein steady-state levels (Figure 3A) and β-galactosidase activities (Figure 3B-3D) of Rad51-NTD-LacZ-NVH and LacZ-NVH in WT, hsp104Δ, new1Δ, doa1Δ, doa4Δ, san1Δ and oaz1Δ yeast cell lines. The protein products encoded by each of the six genes deleted from the latter mutant lines are all functionally relevant to protein homeostasis or prion propagation. Hsp104 is a heat-shock protein with disaggregase activities that disrupts protein aggregation (57, 58). New1 is a translation factor that fine-tunes ribosome recycling and the efficiency of translation termination (53). Doa1 (also called Ufd3) is an ubiquitin- and Cdc48-binding protein with a role in ubiquitin homeostasis and/or protein degradation (59, 60). The doa1Δ mutant exhibits diminished formation of [PSI+] prion (61). Doa4 is a deubiquitinating enzyme required for recycling ubiquitin from proteasome-bound ubiquitinated intermediates (62). The doa4Δ mutant exhibits increased sensitivity to the protein synthesis inhibitor cycloheximide (63). San1 is an ubiquitin-protein ligase that targets highly aggregation-prone proteins (64, 65). Oaz1 (ornithine decarboxylase antizyme) stimulates ubiquitin-independent degradation of Spe1 ornithine decarboxylase by the proteasome (66). We found that the β-galactosidase activities of Rad51-NTD-LacZ-NVH in WT and all six of the gene-knockout strains we examined were 10-29-fold higher than those of LacZ-NVH (Figure 3B). Intriguingly, the β-galactosidase activities of LacZ-NVH in the six gene-knockout mutants are all lower (30-70%) than those in WT (Figure 3C). In contrast, the β-galactosidase activities of Rad51-NTD-LacZ-NVH in WT are either slightly higher or lower than those in the six null mutants (Figure 3D). These results indicate that addition of Rad51-NTD to LacZ-NVH can abrogate the protein homeostasis defects caused by loss of each of these six genes. For example, Rad51-NTD might compensate for the ribosome assembly and translation defects in new1Δ (53), as well as the cycloheximide-hypersensitive phenotype displayed by doa4Δ (63). Accordingly, the β-galactosidase activities of Rad51-NTD-LacZ-NVH in the new1Δ and doa4Δ lines are higher than those in the WT, respectively. In contrast, the β-galactosidase activities of LacZ-NVH in the new1Δ and doa4Δ lines are lower, respectively, than those of WT. Finally, although the doa1Δ mutant is defective in [PSI+] prion formation (61), the steady-state levels of Rad51-NTD-LacZ-NVH in the doa1Δ line are also slightly higher than those in WT.

The expression-promoting function of Rad51-NTD is controlled during protein translation and does not affect ubiquitin-mediated protein degradation.

(A) The steady-state protein levels of Rad51-NTD-LacZ-NVH and LacZ-NVH in WT and six protein homeostasis gene knockout mutants. (B-D) The impact of six protein homeostasis genes on the β-galactosidase activity ratios of Rad51-NTD-LacZ-NVH to LacZ-NVH in WT and the six gene knockout mutants (B). The β-galactosidase activities of LacZ-NVH (C) and Rad51-NTD-LacZ-NVH (D) in WT and the six gene knockout mutants are shown. Asterisks indicate significant differences, with P values calculated using a two-tailed t-test (***, P value <0.001; **, P value <0.01; *, P value <0.05).

The relationship between PEE function, amino acid contents and structural flexibility

Using three Q-rich motifs (i.e., Rad51-NTD, Rad53-SCD1 and Sup35-NPD) as examples, we observed a very strong positive relationship between STQ and STQN amino acid percentages and β-galactosidase activities (Figure 4 and Figure 5). These results are consistent with the notion that, due to high STQ or STQN content, SCDs or Q-rich motifs are intrinsically disordered regions (IDRs) in their native states, rather than adopting stable secondary and/or tertiary structures (31). A common feature of IDRs is their high content of S, T, Q, N, proline (P), glycine (G) and charged amino acids (28-30).

Relative

β-galactosidase (LacZ) activities are correlated with the percentage STQ or STQN amino acid content of three Q-rich motifs. (A) List of N-terminal tags with their respective length, numbers of S/T/Q/N amino acids, overall STQ or STQN percentages, and relative β-galactosidase activities. (B-D) Linear regressions between relative β-galactosidase activities and overall STQ or STQN percentages for Rad51-NTD (B), Rad53-SCD1 (C) and Sup35-PND (D). The coefficients of determination (R2) are indicated for each simple linear regression. (E) The amino acid sequences of wild-type and mutant Rad51-NTD, Rad51-SCD1 and Sup35-PND, respectively.

Alanine scanning mutagenesis of IDRs. The amino acid sequences of WT and mutant IDRs are listed in Table S1. Total protein lysates prepared from yeast cells expressing Rad51-NTD-LacZ-NVH (A), Sup35-PND-LacZ-NVH (B) or Rad53-SCD1-LacZ-NVH (C) were visualized by immunoblotting with anti-V5 antisera. Hsp104 was used as a loading control. Quantitative yeast β-galactosidase (LacZ) assays were carried out as described in Figure 1. Error bars indicate standard deviation between experiments (n = 3). Asterisks indicate significant differences when compared to LacZ-NVH with p values calculated with a two-tailed t-test (**, P value <0.01 and ***, P value <0.001).

It is important to note that the threshold of STQ or STQN content varies in the three cases presented herein (Figure 4B). Thus, the percentage of STQ or STQN residues is not likely the only factor contributing to protein expression levels. Since G, P and glutamate (E) are enriched by >10% in Rad51-NTD, Rad53-SCD1 and Sup35-NPD, these three amino acids may also contribute to the PEE activities and structural flexibility of these three Q-rich motifs.

Functionally, IDRs are key components of subcellular machineries and signaling pathways because they have the potential to associate with many partners due to their ability to adopt multiple metastable conformations. Many IDRs are regulated by alternative splicing and post-translational modifications, and some are involved in the formation of various membraneless organelles or proteinaceous condensates via intracellular liquid-liquid phase separation (67, 68). Since IDRs can endow proteins with structural and functional plasticity, we hypothesized that Q-rich motifs (e.g., SCD, polyQ and polyQ/N) represent useful toolkits for creating new diversity during protein evolution.

Comparative GO enrichment analyses of SCDs and polyX motifs in different eukaryotes

To verify our hypothesis, we designed two JavaScript software programs (ASFinder-SCD and ASFinder-polyX) for proteome-wide searches of amino acid sequences that contain ≥3 S/T-Q motifs within a stretch of ≤100 residues (32) and the polyX motifs of 20 different amino acids, respectively. In the latter case, diverse thresholds have been used in different studies or databases to define and detect polyX motifs. We applied the lowest threshold (≥4/7), i.e., a minimum number of four identical X amino acid residues in a localized region of seven amino acid residues (69-72) (Figures 6-9 and Dataset DS1-DS29). We searched and compared the near-complete proteomes of 27 different eukaryotes (Table 4), including the budding yeast S. cerevisiae, three pathogenic species of Candida, three filamentous ascomycete fungi (Neurospora crassa, Magnaporthe oryzae and Trichoderma reesei), three basidiomycete fungi (Cryptococcus neoformans, Ustilago maydis and Taiwanofungus camphoratus), the slim mold Dictyostelium discoideum, the malaria-causing unicellular protozoan parasite Plasmodium falciparum, six unicellular ciliates (Tetrahymena thermophila, Paramecium tetraurelia, Oxytricha trifallax, Stylonychia lemnae, Pseudocohnilembus persalinus and Stentor coeruleus), the fly Drosophila melanogaster, the mosquito Aedes aegypti, the nematode Caenorhabditis elegans, the zebrafish Danio rerio, the mouse Mus musculus, Homo sapiens, the higher plant Arabidopsis thaliana, and the single-celled green alga Chlamydomonas reinhardtii. The Benchmarking Universal Single-Copy Ortholog (BUSCO) scores of the near-universal single-copy gene orthologs of all 27 proteomes are 92.4-100% (Table 4). Genome or protein matrix scores >95% for model organisms are generally deemed complete reference genomes or proteomes (73).

Percentages of proteins with different numbers of SCDs, and polyQ, polyQ/N or polyN tracts in 37 different eukaryotes.

It was reported previously that SCDs are overrepresented in the yeast and human proteomes (32, 74), and that polyX prevalence differs among species (43, 75-77). Our results reveal that the percentages of SCD proteins in the near-complete proteomes of 21 non-ciliate species and 6 ciliates range from 8.0% in P. falciparum, 13.9% in H. sapiens, 16.8% in S. cerevisiae, 24.2% in U. maydis, to a maximum of 58.0% in O. trifallax (Figure 6 and Dataset DS2). Among the 6050 proteins in the most recently modified S. cerevisiae reference proteome (https://www.uniprot.org/proteomes/UP000002311), we identified 1016 SCD-hosting proteins (Dataset DS4), including all 436 SCD-harboring proteins previously revealed by ScanProsite (32). ScanProsite is a publicly available database of protein families, domains and motifs (78).

The percentages of polyQ proteins in the near-complete proteomes of the 21 non-ciliate species and 6 ciliates range from 0.6% (S. coeruleus), 2.1% (P. falciparum), 4.1% (S. cerevisiae), 6.5% (H. sapiens), 21.8% (D. melanogaster), 24.5% (D. discoideum) to a maximum of 47.2% in Pseudocohnilembus persalinus. Those of polyQ/N proteins range from 9.5% in T. camphoratus, 11.9% in H. sapiens, 28.2% in S. cerevisiae, 62.5% in D discoideum, 63.7% in P. falciparum, to a maximum of 79.4% in P. persalinus. Percentages of polyN proteins range from 0.9% in T. camphoratus, 1.0% in H. sapiens, 1.1% in M. musculus, 3.7% in S. coeruleus, 49.7% in D. discoideum, to a maximum of 54.0% in P falciparum (Dataset DS2). Compared to other species, D. melanogaster and A. aegypti possess higher percentages of polyH proteins; P. falciparum has higher percentages of poly-phenylalanine (polyF) and polyY proteins; and C. reinhardtii has higher percentages of polyA, polyP, polyG and polyH proteins.

Notably, average usage of a given amino acid (X) is not always positively correlated with the percentages of polyX proteins in different species. For instance, proteome-wide Q usages are 2.8% in P. falciparum, 3.7% in S. coeruleus, 5.4% in D. melanogaster and 10.9% in P. persalinus. The proteome-wide usage of F in P. falciparum (4.4%) is not significantly different from that of other species, ranging from 2.0% in C. reinhardtii to 5.2% in T. thermophila. Average usages of P (7.7%), A (17.1%) and G (11.5%) in C. reinhardtii, as well as the average usages of N (14.4%) and Y (5.7%) in P. falciparum, are higher than those in 6 ciliate and the 20 other non-ciliate species, respectively (Figure 6 and Dataset DS1). These results indicate that distinct evolutionary strategies have evolved to enable specific polyX motifs to accumulate in different species.

Next, we performed comparative GO enrichment analyses using information on the functions of genes provided by GO knowledgebase (http://geneontology.org). Our results provide two intriguing insights into evolutionary biology. First, SCDs (Table 2) and polyQ/N motifs (Table 6) exist in many proteins involved in various evolutionarily conserved processes (Dataset DS4-DS29). It was reported previously that many SCD proteins in yeast and human are neither targets of ATR or ATM nor are related to pathways of functions relevant to the DDR or DNA replication, including telophase, cytokinesis, protein transport, endocytosis and cytoskeleton (32). Second, SCDs (Table 5) and polyQ/N motifs (Table 6) are overrepresented in the specialized biological processes of different species. Of particular note is an overrepresentation of conserved SCD proteins (74) and polyQ/N proteins (this study) in pathways related to vertebrate neural development and neurodegeneration. For instance, human CTTNBP2 (1663 amino acid residues) is a neuron-specific F-actin associated SCD-protein that is involved in the formation and maintenance of dendritic spines and it is associated with autism spectrum disorders (79, 80). Human CTTNBP2 possesses ten S/T-Q motifs (T466Q, T493Q, S1580Q, S553Q, S634Q, S994Q, S1392Q, S1580Q, T1621Q and S1624Q). Mouse CTTNBP2 has 630 amino acid residues, four S/T-Q motifs (S419Q, T463Q, S550Q, S624Q), and it shares high amino acid identity to the N-terminus (1-640 amino acid residues) of human CTTNBP2. Using IUPred2A (https://iupred2a.elte.hu/plot_new), a web-server for identifying disordered protein regions (81), we found that both human CTTNBP2 (220-1633 residues) and mouse CTTNBP2 (220-630 residues) are Q/N-rich IDRs with high percentages of S, T, Q, N, G, P, R and K. We reported recently that mouse CTTNBP2 forms self-assembled condensates through its C-terminal IDR and it facilitates co-condensation of an abundant excitatory postsynaptic scaffold protein SHANK3 at dendritic spines in a Zn2+-dependent manner (82). Our results also reveal that SCDs, as well as polyQ, polyQ/N, polyS and polyP tracts, are overrepresented in the nervous system of D. melanogaster (Tables 5 and 6). The numbers (and percentages) of SCD-, polyQ-, polyQ/N-, polyS- and polyP-containing proteins, respectively, in different nerve-related biological processes are: regulation of axonogenesis [101 (87%), 93 (80%), 98 (84%), 103 (89%) and 100 (86%)], neuron development [97 (80%), 99 (82%), 102 (84%), 103 (85%) and 88 (73%)], axon guidance [199 (53%), 209 (56%), 278 (74%), 280 (75%), and 186 (50%)], mushroom body development [140 (66%), 142 (67%), 167 (79%), 161 (76%) and 109 (51%)], dendritic morphogenesis [96 (53%), 95 (52%), 132 (73%), 131 (72%) and 61 (34%)], and peripheral nervous system development [94 (73%), 100 (78%), 111 (86%), 106 (82%) and 90 (70%)] (Table 5, Table 6 and Dataset DS22).

We also found SCDs, polyQ/N tracts and a few other structurally flexible polyX tracts that are overrepresented in specialized biological processes of various species. First, in S. cerevisiae, 92 proteins are involved in RNA-mediated transposition (GO ID: 32197), among which 82 and 88 are SCD and polyQ/N proteins, respectively (Table 5, Table 6 and Dataset DS4). The 82 SCD (also polyQ/N) proteins are all structural components of the virus-like particles (VLPs), which form the shell that encapsulates the Ty1 (Transposon-yeast 1) and Ty2 retrotransposon dimeric RNA genomes. These capsid proteins also possess nucleocapsid-like chaperone activity, promoting primer tRNA(i)-Met annealing to the multipartite primer-binding site, dimerization of Ty RNA, and initiation of reverse transcription (83). Another example is the developmental and sexual cycles of D discoideum, which are induced by starvation and/or other stimuli. During the developmental cycle, these amoebae secrete small molecules (e.g., cyclic adenosine monophosphate or cAMP) to induce intercellular communication (chemotaxis), ultimately resulting in aggregation of hundreds to hundreds of thousands of amoebae into multicellular fruiting bodies (sorocarps). These sorocaps consist of ∼80% asexual spores and ∼20% dead stalk cells. During the sexual cycle, two amoebae of complementary mating types fuse to form a zygote. The zygote and the surrounding amoebae form a thick multilayered structure called the macrocyst, representing the site of meiosis and interhomolog DNA recombination, enabling the generation of hundreds of recombinant haploid amoebae (84, 85). We found that SCDs (Table 5) and polyQ/N tracts (Table 6) are overrepresented in many proteins involved in a diversity of biological processes during the D. discoideum development and sexual cycles, including chemotaxis to cAMP, amoebae aggregation, sorocarp morphogenesis, sorocarp development, and sexual reproduction (Dataset DS14). A third example is the human malaria-causing parasite P. falciparum. Infection in humans begins upon being bitten by an infected female Anopheles mosquito. After entering the human bloodstream, sporozoites from the mosquito’s salivary gland quickly invade hepatocytes and undergo asexual multiplication, leading to tens of thousands of merozoites bursting from the liver cells. These merozoites then invade red blood cells (erythrocytes) and undergo an additional round of multiplication. The mature merozoites can change the surface properties of infected red blood cells, causing them to stick to blood vessels (a process called cytoadherence), resulting in obstructed microcirculation and multiple organ dysfunction. The success of P. falciparum in establishing persistent infections is mainly attributable to immune evasion through antigenic variation (86). Our GO enrichment results reveal that SCDs, as well as polyQ/N, polyN, polyS, polyT, polyD, polyG and polyK tracts, are overrepresented in many of these specialized infection processes, including antigen variation, modulation of host erythrocyte aggregation, cytoadherence to the microvasculature, and cell-cell adhesion (Table 5, Table 6 and Dataset DS15). For example, the numbers of proteins involved in antigen variation are 38 (SCDs), 82 (polyQ/N), 37 (polyN), 51 (polyS), 49 (polyT), 42 (polP), 69 (polyG), 40 (polyD), 68 (polyE) and 144 (polyK), respectively. These examples are consistent with the notion that S, T, Q, N, P, G and charged amino acids are used more frequently in IDRs (28-30).

Massive expansion of Q-rich motifs in ciliates having two reassigned stop codons (UAAQ and UAGQ)

The most striking finding in our study is that, due to their usage of the two noncanonical codons (UAAQ and UAGQ), Q (but not S, T or N) is used more frequently in five free-moving unicellular ciliates (T. thermophila, P. tetraurelia, O. trifallax, S. lemnae and P. persalinus) than in the sessile ciliate S. coeruleus or in any of the 21 non-ciliate species we examined herein (Figures 6 and Dataset DS1). We refer hereafter to these five free-moving unicellular ciliates as ‘group I’ ciliates. The proteome-wide Q and N contents, as well as the percentages of SCD, polyQ, polyQ/N and polyN proteins, respectively, in the different ciliates are: T. thermophila (9.6%, 9.2%, 50.8%, 33.0%, 72.6%, 25.2%), P. tetraurelia (9.1%, 6.9%, 41.1%, 27.1%, 62.9% and 8.7%), O. trifallax (9.0%, 7.9%, 58.0%, 36.7%, 72.3%, 22.0%), S. lemnae (8.4%, 7.5%, 49.2%, 28,1%, 65.3%, 17.4%), P. persalinus (10.9%, 9.8%, 47.7%, 47.2%, 79.5%, 36.3%), and S. coeruleus (3.7%, 6.3%, 13.9%, 0.6%, 18.7% and 3.7%] (Figure 6-9, Dataset DS2 and DS3).

Our GO enrichment results also reveal that the percentages of SCD, polyQ and polyQ/N (but not polyN) proteins in many biological processes of the five group I ciliates are significantly higher than those in S. coeruleus. For example, the numbers (and percentages) of SCD, polyQ, polyQ/N and polyN proteins involved, respectively, in DNA repair (GO ID: 06281) are: T. thermophila [28 (61%), 18 (39%), 38 (83%) and 13 (28%)] (DS16); P. tetraurelia [27 (44%), 21 (34%), 49 (79%) and 3 (5%)] (DS17); O. trifallax [50 (61%), 35 (43%), 62 (76%) and 22 (27%)] (DS18); S. lemnae [32 (53%), 21 (35%), 44 (73%) and 19 (17%)] (DS19); P. persalinus [19 (53%), 19 (53%), 31 (86%) and 10 (28%)] (DS20); and S. coeruleus [7 (17%), 1 (2%), 10 (24%) and 2 (5%)] (DS22). The respective numbers (and percentages) of SCD, polyQ, polyQ/N and polyN proteins involved in microtubule-based movement (GO ID: 07018) are: T. thermophila [93 (85%), 69 (63%), 107 (97%) and 63 (57%)] (DS16); P. tetraurelia [155 (69%), 109 (48%), 195 (86%) and 47 (21%)] (DS17); O. trifallax [99 (78%), 85 (66%), 121 (95%) and 47 (37%)] (DS18); S. lemnae [84 (80%), 66 (63%), 100 (95%) and 50 (48%)] (DS19); P. persalinus [65 (72%), 63 (70%), 85 (94%) and 41 (46%)] (DS20); and S. coeruleus [78 (34%), 5 (2%), 84 (37%) and 13 (6%)] (DS21).

Compared to 20 non-ciliate species, all five group I ciliates and the sessile S. coeruleus possess proteins involved in peptidyl-glutamic acid modification (GO ID:18200), with respective numbers (and percentages) of SCD, polyQ, polyQ/N and polyN proteins being: T. thermophila [38 (78%), 31 (63%), 47 (96%) and 23(47%)] (Dataset DS16); P. tetraurelia [42 (60%), 23 (33%), 58 (83%) and 5 (7%)] (Dataset DS17); O. trifallax [81 (91%), 63 (71%), 84 (94%) and 50 (56%)] (Dataset DS18); S. lemnae [63 (86%), 54 (74%), 68 (93%) and 34 (47%)] (Dataset DS19); P. persalinus [38 (58%), 37 (57%), 55 (85%) and 28 (43%)] (Dataset DS20); and S. coeruleus [30 (23%), 0 (0%), 50 (38%) and 23 (17%)] (Dataset DS21).

T. thermophila has 58 proteins involved in the xylan catabolic process (GO ID: 45493) (Table 7), of which 56 (97%) are SCDs, 55 (95%) are polyQ proteins, 58 (100%) are polyQ/N proteins, and 49 (84%) are polyN proteins (Dataset DS16). Using the NCBI BLASTP search tool with an expect value (E-value) ≤ 10e-5 to search for homologs of these 58 proteins among all other 16 ciliates analyzed in this study, only 144 proteins with amino acid identity <60% and a raw alignment score of <150 were found (Dataset DS30).

In T. thermophila, there are 124 annotated proteins involved in various events during meiosis, including meiosis initiation and regulation, homologous pairing, formation of DNA DSBs, formation of CO and chiasma (a structure that forms between a pair of homologous chromosomes by CO recombination and physically links the homologous chromosomes during meiosis), as well as meiotic nuclear divisions (DS31) (see review in (8)). Among the 124 T. thermophila meiotic proteins, we identified 85 SCD proteins, 54 polyQ proteins, 106 polyQ/N proteins, 32 polyN proteins, 32 polyS proteins and 48 polyK proteins, respectively. Notably, there are 48 and 59 meiotic proteins that contain ≥4 SCDs and/or ≥4 polyQ/N tracts, respectively. For instance, DPL2, CYC2, CYC17 and ATR1 each contains 15, 6, 8 and 5 SCDs, 2, 0, 1 and 2 polyQ tracts, as well as 11, 4, 4 and 5 polyQ/N tracts, respectively. Pars11, a chromatin-associated protein required for inducing and limiting Spo11-induced DSBs, has 14 SCDs, 2 polyQ tracts and 6 polyQ/N tracts. Spo11-induced DSBs promote ATR1-dependent Pars11 phosphorylation and its removal from chromatin (13). Many T. thermophila meiotic DSB repair proteins also harbor several SCDs, polyQ tracts and/or polyQ/N tracts, including MSH4, MSH5, SGS1, FANCM, REC8, and ZPH3, among others. Several T. thermophila proteins involved in editing and rearrangement of the MIC genome also harbor multiple Q-rich motifs. PDD1 (programmed DNA degradation 1), a conjugation-specific HP1-like protein, has 13 SCDs, 3 polyQ tracts and 6 polyQ/N tracts. Mutations in the chromodomain or the chromoshadow domain of PDD1 were found previously to elicit PDD1 mislocalization, prevented histone H3 dimethylation on K9, abolished removal of internal eliminated sequences (IES), and/or resulted in the production of inviable progeny (87). DCL1 (Dicer-like 1) has 5 SCDs, 1 polyQ tract and 8 polyQ/N tracts. DCL1 is required for processing the MIC transcripts to siRNA-like scan (scn) RNAs, as well as for methylation of histone H3 at Lys 9. This latter modification occurs specifically on sequences (IESs) to be eliminated (88). GIW1 (gentlemen-in-waiting 1) physically directs a mature Argonaute-siRNA complex to the MIC nucleus, thus promoting programmed IES elimination (89). GIW1 has 9 SCDs and 2 polyQ/N tracts (Dataset DS31). Using IUPred2A (https://iupred2a.elte.hu/plot_new), we found that the Q-rich motifs in all these meiotic proteins are intrinsically disordered. Accordingly, it is of interest to speculate that, like the C-terminal IDR of mammalian CTTNBP2, the Q-rich motifs in T. thermophila meiotic proteins might form tunable proteinaceous condensates to regulate assembly and disassembly of the “crescents” in ciliate MICs.

Finally, we performed comparative GO analyses for the SCD and polyX proteins encoded by the transcriptomes of 11 different ciliate species. These transcripts were originally generated as part of the Marine Microbial Eukaryote Transcriptome Sequencing Project (MMETSP) (90), which were then reassembled and reannotated by Brown and colleagues (91). All transcripts are publicly available at Zendo (https://zenodo.org/record/1212585#.Y79zoi2l3PA) (91). We applied TransDecoder (Find Coding Regions Within Transcripts; https://github.com/TransDecoder/TransDecoder/wiki) to identify candidate coding regions within the transcript sequences. Five of those eleven ciliates have reassigned UAAQ and UAGQ codons (hereafter termed “group II ciliates”), i.e., Aristerostoma spp., Favella ehrenbergii, Pseudokeronopsis spp., Strombidium inclinatum, and Uronema spp.. Like sessile S. coeruleus, group III ciliates (Climacostomum virens, Litonotus pictus and Protocruzia adherens) possess the standard genetic codes. Group IV ciliates encompass Favella ehrenbergii, Blepharisma japonicum and Euplotes focardii, each of which utilizes the reassigned codons UAAY, UAGY, UGAW and UGAC, respectively (Table 1). The proteins encoded by these MMESTP transcripts are unlikely to represent the entire protein complement of all 11 ciliate species because 1) many MMETSP transcripts are not intact (i.e., broken mRNAs) and thus encode incomplete protein sequences, and 2) except for C. virens (94.7%), the BUSCO protein scores of these MMESTP transcripts only range from 52.6% to 89.9% (Table 1). Nevertheless, our results indicate that Q is used more frequently in group I and group II ciliates than in group III and group IV ciliates or in a further 20 non-ciliate species (Figure 3 and DS1). Accordingly, proportions of SCD, polyQ and polyQ/N proteins in all group I and group II ciliates are higher than they are in the three group III ciliates (except L. pictus) and the three group IV ciliates, respectively (Figure 4-6). Since N is used more frequently in L. pictus than in group II ciliates, as well as in the other group III and group IV ciliates, it has higher percentages of polyN and polyQ/N proteins (Figures 6-9 and Dataset DS1).

Our data also indicates that Y, W or C are not used more frequently in the three group IV ciliates than in the other 14 ciliate or 20 non-ciliate species (Figure 7 and Dataset DS1). Reassignments of stop codons to Y, W or C also do not result in higher percentages of polyY, polyW or polyC proteins in the three group IV ciliates, respectively (Figure 9).

Proteome-wide contents of 20 different amino acids in 37 different eukaryotes.

Percentages of proteins with indicated polyQ and polyQ/N tracts in 37 different eukaryotes.

Percentages of proteins with indicated polyX motifs in 37 different eukaryotes.

Discussion

We present two unexpected results in this report. First, the Q-rich motifs of several yeast proteins (Rad51-NTD, Rad53-SCD1, Hop1-SCD, Sml1-NTD, Sup35-PND, Ure2-UPD and New1-NPD) all exhibit autonomous PEE activities. These structurally flexible Q-rich motifs have useful potential for applications in both basic research (e.g., synthetic biology) and biotechnology. It is of interest to investigate further how these Q-rich motifs exert this PEE function in yeast and whether Q-rich motifs in other eukaryotes also possess similar PEE activities. Second, reassignment of stop codons to Q in the group I and group II ciliates significantly increase proteome-wide Q usage, leading to massive expansion of structurally flexible or even intrinsically disordered Q-rich motifs. In contrast, reassignments of stop codons to Y, W or C do not result in higher usages of these three amino acid residues nor higher percentages of W-, Y- or C-rich proteins in the three group IV ciliates, respectively. These results are consistent with the notion that, unlike for Q, the Y, W and C residues are not common in IDRs (28-30).

Due to their structural flexibility, Q-rich motifs can endow proteins with structural and functional plasticity. Based on our findings from this study, we suggest that Q-rich motifs are useful toolkits for generating novel diversity during protein evolution, including by enabling greater protein expression, protein-protein interactions, posttranslational modifications, increased solubility, and tunable stability, among other important traits. This speculation may explain three intriguing phenomena. First, due to higher Q usage, many proteins involved in evolutionarily conserved biological processes in group I and group II ciliates display more diverse amino acid sequences than the respective proteins in other ciliate or non-ciliate species. Accordingly, it is sometimes difficult to identify authentic protein homologs among different ciliates, particularly for group I and group II ciliates. We highlight the example of the 58 proteins involved in xylan catabolysis in T. thermophila (Dataset DS16). Second, our GO enrichment results reveal that Q-rich motifs or a few other structurally flexible polyX tracts prevail in proteins involved in specialized biological processes, including S. cerevisiae RNA-mediated transposition, C. albicans filamentous growth, ciliate peptidyl-glutamic acid modification, T. thermophila meiosis and xylan catabolism, D. discoideum sorocarp (fruiting body) development and sexual reproduction, P. falciparum blood cell infection and antigen variation, and development of the D. melanogaster nervous system. In theory, structurally flexible Q-rich motifs might form various membraneless organelles or proteinaceous condensates via intracellular liquid-liquid phase separation in tunable manners, including protein posttranslational modification, protein-protein interaction, protein-ligand binding, et al., (67, 68). A typical example is that the C-termini of mouse CTTNBP2 facilitates co-condensation of CTTNBP2 with SHANK3 at the postsynaptic density in a Zn2+-dependent manner (82). Third, the Borg genomes of anaerobic methane-oxidizing archaea are huge, linear extrachromosomal elements. A striking feature of Borgs is pervasive tandem direct repeat (TR) regions. While we were preparing this manuscript, it was reported that TRs in ORFs are under very strong selective pressure, leading to perfect amino acid TRs (aaTRs) that are commonly IDRs. aaTRs particularly often contain disorder-promoting amino acids, including Q, N, P, T, E, K, V, D and S, respectively (92). Accordingly, distinct evolutionary strategies have evolved in different species to evolve protein regions that are structurally flexible or even intrinsically disordered. Further investigations are needed to determine if liquid-liquid phase separation prevails in ciliates with reassigned TAAQ and TAGQ codons and/or in the specialized biological processes in various species we have described herein.

Conclusions

One of the most interesting questions in genome diversity is why many ciliates reassign their nuclear stop codons into amino acids, e.g., glutamine (Q), tyrosine (Y), tryptophan (W) or cysteine (C). However, the impacts of such genome-wide alternations are not well understood. We show that glutamine (Q) is used more frequently in all 10 ciliate species possessing reassigned TAAQ and TAGQ codons than in other ciliates and non-ciliate species. The consequence of this preponderance of Q is massive expansion of proteins harboring three structurally flexible or even intrinsically disordered Q-rich motifs. Since Q-rich motifs can endow proteins with structural and functional plasticity, we also show that Q-rich motifs prevail in proteins involved in several species-specific or even phylum-specific biological processes. Our results indicate that Q-rich motifs are useful toolkits for evolutionary novelty.

Methods

All plasmids, yeast strains and PCR primers used in this study are listed in the Tables 1-3, respectively. Guinea pig antisera against Rad51, and rabbit antisera against phosphorylated Rad51-S12Q, phosphorylated Rad51-S30Q, and phosphorylated Hop1-T318Q were described previously (27, 40). The mouse anti-V5 antibody was purchased from BioRad (CA, USA). The rabbit anti-Hsp104 antiserum was kindly provided by Chung Wang (Academia Sinica, Taiwan). Rabbit antisera against phosphorylated Sup35-S17Q were raised using the synthetic phosphopeptide N12YQQYS(P)QNGNQQQGNNR28 as an antigen, where S(P) is phosphorylated serine. Phosphopeptide synthesis and animal immunization were conducted by LTK BioLaboratories, Taiwan. Western blotting analyses were performed as described previously (27, 40). Quantitative β-galactosidase activity assays were carried out as previously described (27, 93). The ASFinder-SCD and ASFinder-polyX software are publicly available at Github (https://github.com/tfwangasimb/AS-SCD-finder). The sources of proteome and transcript datasets are described in Table 4. The proteome-wide amino acid contents, the numbers of protein with different types of Q-rich motifs and polyN, as well as the SCD and polyX proteins in all 37 different eukaryotes are listed in Datasets DS1-DS3, respectively. The Gene Ontology (GO) enrichment analyses were performed using publicly available data in the GO Resource (http://geneontology.org). The GO identities (ID) of different biological processes and the names of all SCD and polyX proteins in the 24 near-completed eukaryotic proteomes are listed in Dataset DS4-DS29, respectively.

Supporting information

eLife Assessment

This study presents potentially valuable results on glutamine-rich motifs in relation to protein expression and alternative genetic codes. The author's interpretation of the results is so far only supported by incomplete evidence, due to a lack of acknowledgment of alternative explanations, missing controls and statistical analysis and writing unclear to non experts in the field. These shortcomings could be at least partially overcome by additional experiments, thorough rewriting, or both.

Reviewer #1 (Public Review):

Summary

This work contains 3 sections. The first section describes how protein domains with SQ motifs can increase the abundance of a lacZ reporter in yeast. The authors call this phenomenon autonomous protein expression-enhancing activity, and this finding is well supported. The authors show evidence that this increase in protein abundance and enzymatic activity is not due to changes in plasmid copy number or mRNA abundance, and that this phenomenon is not affected by mutants in translational quality control. It was not completely clear whether the increased protein abundance is due to increased translation or to increased protein stability.

In section 2, the authors performed mutagenesis of three N-terminal domains to study how protein sequence changes protein stability and enzymatic activity of the fusions. These data are very interesting, but this section needs more interpretation. It is not clear if the effect is due to the number of S/T/Q/N amino acids or due to the number of phosphorylation sites.

In section 3, the authors undertake an extensive computational analysis of amino acid runs in 27 species. Many aspects of this section are fascinating to an expert reader. They identify regions with poly-X tracks. These data were not normalized correctly: I think that a null expectation for how often poly-X track occur should be built for each species based on the underlying prevalence of amino acids in that species. As a result, I believe that the claim is not well supported by the data.

Strengths

This work is about an interesting topic and contains stimulating bioinformatics analysis. The first two sections, where the authors investigate how S/T/Q/N abundance modulates protein expression level, is well supported by the data. The bioinformatics analysis of Q abundance in ciliate proteomes is fascinating. There are some ciliates that have repurposed stop codons to code for Q. The authors find that in these proteomes, Q-runs are greatly expanded. They offer interesting speculations on how this expansion might impact protein function.

Weakness

At this time, the manuscript is disorganized and difficult to read. An expert in the field, who will not be distracted by the disorganization, will find some very interesting results included. In particular, the order of the introduction does not match the rest of the paper.

In the first and second sections, where the authors investigate how S/T/Q/N abundance modulates protein expression levels, it is unclear if the effect is due to the number of phosphorylation sites or the number of S/T/Q/N residues. The authors also do not discuss if the N-end rule for protein stability applies to the lacZ reporter or the fusion proteins.

The most interesting part of the paper is an exploration of S/T/Q/N-rich regions and other repetitive AA runs in 27 proteomes, particularly ciliates. However, this analysis is missing a critical control that makes it nearly impossible to evaluate the importance of the findings. The authors find the abundance of different amino acid runs in various proteomes. They also report the background abundance of each amino acid. They do not use this background abundance to normalize the runs of amino acids to create a null expectation from each proteome. For example, it has been clear for some time (Ruff, 2017; Ruff et al., 2016) that Drosophila contains a very high background of Q's in the proteome and it is necessary to control for this background abundance when finding runs of Q's. The authors could easily address this problem with the data and analysis they have already collected. However, at this time, without this normalization, I am hesitant to trust the lists of proteins with long runs of amino acid and the ensuing GO enrichment analysis.

Ruff KM. 2017. Washington University in St.

Ruff KM, Holehouse AS, Richardson MGO, Pappu RV. 2016. Proteomic and Biophysical Analysis of Polar Tracts. Biophys J 110:556a.

Reviewer #2 (Public Review):

Summary:

This study seeks to understand the connection between protein sequence and function in disordered regions enriched in polar amino acids (specifically Q, N, S and T). While the authors suggest that specific motifs facilitate protein-enhancing activities, their findings are correlative, and the evidence is incomplete. Similarly, the authors propose that the re-assignment of stop codons to glutamine-encoding codons underlies the greater user of glutamine in a subset of ciliates, but again, the conclusions here are, at best, correlative. The authors perform extensive bioinformatic analysis, with detailed (albeit somewhat ad hoc) discussion on a number of proteins. Overall, the results presented here are interesting, but are unable to exclude competing hypotheses.

Strengths:

Following up on previous work, the authors wish to uncover a mechanism associated with poly-Q and SCD motifs explaining proposed protein expression-enhancing activities. They note that these motifs often occur IDRs and hypothesize that structural plasticity could be capitalized upon as a mechanism of diversification in evolution. To investigate this further, they employ bioinformatics to investigate the sequence features of proteomes of 27 eukaryotes. They deepen their sequence space exploration uncovering sub-phylum-specific features associated with species in which a stop-codon substitution has occurred. The authors propose this stop-codon substitution underlies an expansion of ploy-Q repeats and increased glutamine distribution.

Weaknesses:

The preprint provides extensive, detailed, and entirely unnecessary background information throughout, hampering reading and making it difficult to understand the ideas being proposed.

The introduction provides a large amount of detailed background that appears entirely irrelevant for the paper. Many places detailed discussions on specific proteins that are likely of interest to the authors occur, yet without context, this does not enhance the paper for the reader.

The paper uses many unnecessary, new, or redefined acronyms which makes reading difficult. As examples: (1) Prion forming domains (PFDs). Do the authors mean prion-like domains (PLDs), an established term with an empirical definition from the PLAAC algorithm? If yes, they should say this. If not, they must define what a prion-forming domain is formally. (2) SCD is already an acronym in the IDP field (meaning sequence charge decoration) - the authors should avoid this as their chosen acronym for Serine(S) / threonine (T)-glutamine (Q) cluster domains. Moreover, do we really need another acronym here (we do not). (3) Protein expression-enhancing (PEE) - just say expression-enhancing, there is no need for an acronym here.

The results suggest autonomous protein expression-enhancing activities of regions of multiple proteins containing Q-rich and SCD motifs. Their definition of expression-enhancing activities is vague and the evidence they provide to support the claim is weak. While their previous work may support their claim with more evidence, it should be explained in more detail. The assay they choose is a fusion reporter measuring beta-galactosidase activity and tracking expression levels. Given the presented data they have shown that they can drive the expression of their reporters and that beta gal remains active, in addition to the increase in expression of fusion reporter during the stress response. They have not detailed what their control and mock treatment is, which makes complete understanding of their experimental approach difficult. Furthermore, their nuclear localization signal on the tag could be influencing the degradation kinetics or sequestering the reporter, leading to its accumulation and the appearance of enhanced expression. Their evidence refuting ubiquitin-mediated degradation does not have a convincing control.

Based on the experimental results, the authors then go on to perform bioinformatic analysis of SCD proteins and polyX proteins. Unfortunately, there is no clear hypothesis for what is being tested; there is a vague sense of investigating polyX/SCD regions, but I did not find the connection between the first and section compelling (especially given polar-rich regions have been shown to engage in many different functions). As such, this bioinformatic analysis largely presents as many lists of percentages without any meaningful interpretation. The bioinformatics analysis lacks any kind of rigorous statistical tests, making it difficult to evaluate the conclusions drawn.

The methods section is severely lacking. Specifically, many of the methods require the reader to read many other papers. While referencing prior work is of course, important, the authors should ensure the methods in this paper provide the details needed to allow a reader to evaluate the work being presented. As it stands, this is not the case.

Overall, my major concern with this work is that the authors make two central claims in this paper (as per the Discussion).

The authors claim that Q-rich motifs enhance protein expression. The implication here is that Q-rich motif IDRs are special, but this is not tested. As such, they cannot exclude the competing hypothesis ("N-terminal disordered regions enhance expression"). The authors also do not explore the possibility that this effect is in part/entirely driven by mRNA-level effects (see Verma Na Comms 2019). As such, while these observations are interesting, they feel preliminary and, in my opinion, cannot be used to draw hard conclusions on how N-terminal IDR sequence features influence protein expression. This does not mean the authors are necessarily wrong, but from the data presented here, I do not believe strong conclusions can be drawn.

That re-assignment of stop codons to Q increases proteome-wide Q usage. I was unable to understand what result led the authors to this conclusion. My reading of the results is that a subset of ciliates has re-assigned UAA and UAG from the stop codon to Q. Those ciliates have more polyQ-containing proteins. However, they also have more polyN-containing proteins and proteins enriched in S/T-Q clusters. Surely if this were a stop-codon-dependent effect, we'd ONLY see an enhancement in Q-richness, not a corresponding enhancement in all polar-rich IDR frequencies? It seems the better working hypothesis is that free-floating climate proteomes are enriched in polar amino acids compared to sessile ciliates. Regardless, the absence of any kind of statistical analysis makes it hard to draw strong conclusions here.

Author Response

The following is the authors’ response to the original reviews.

eLife assessment

This study presents potentially valuable results on glutamine-rich motifs in relation to protein expression and alternative genetic codes. The author's interpretation of the results is so far only supported by incomplete evidence, due to a lack of acknowledgment of alternative explanations, missing controls and statistical analysis and writing unclear to non experts in the field. These shortcomings could be at least partially overcome by additional experiments, thorough rewriting, or both.

We thank both the Reviewing Editor and Senior Editor for handling this manuscript.

Based on your suggestions, we have provided controls, performed statistical analysis, and rewrote our manuscript. The revised manuscript is significantly improved and more accessible to non-experts in the field.

Reviewer #1 (Public Review):

Summary

This work contains 3 sections. The first section describes how protein domains with SQ motifs can increase the abundance of a lacZ reporter in yeast. The authors call this phenomenon autonomous protein expression-enhancing activity, and this finding is well supported. The authors show evidence that this increase in protein abundance and enzymatic activity is not due to changes in plasmid copy number or mRNA abundance, and that this phenomenon is not affected by mutants in translational quality control. It was not completely clear whether the increased protein abundance is due to increased translation or to increased protein stability.

In section 2, the authors performed mutagenesis of three N-terminal domains to study how protein sequence changes protein stability and enzymatic activity of the fusions. These data are very interesting, but this section needs more interpretation. It is not clear if the effect is due to the number of S/T/Q/N amino acids or due to the number of phosphorylation sites.

In section 3, the authors undertake an extensive computational analysis of amino acid runs in 27 species. Many aspects of this section are fascinating to an expert reader. They identify regions with poly-X tracks. These data were not normalized correctly: I think that a null expectation for how often poly-X track occur should be built for each species based on the underlying prevalence of amino acids in that species. As a result, I believe that the claim is not well supported by the data.

Strengths

This work is about an interesting topic and contains stimulating bioinformatics analysis. The first two sections, where the authors investigate how S/T/Q/N abundance modulates protein expression level, is well supported by the data. The bioinformatics analysis of Q abundance in ciliate proteomes is fascinating. There are some ciliates that have repurposed stop codons to code for Q. The authors find that in these proteomes, Q-runs are greatly expanded. They offer interesting speculations on how this expansion might impact protein function.

Weakness

At this time, the manuscript is disorganized and difficult to read. An expert in the field, who will not be distracted by the disorganization, will find some very interesting results included. In particular, the order of the introduction does not match the rest of the paper.

In the first and second sections, where the authors investigate how S/T/Q/N abundance modulates protein expression levels, it is unclear if the effect is due to the number of phosphorylation sites or the number of S/T/Q/N residues.

There are three reasons why the number of phosphorylation sites in the Q-rich motifs is not relevant to their autonomous protein expression-enhancing (PEE) activities:

First, we have reported previously that phosphorylation-defective Rad51-NTD (Rad51-3SA) and wild-type Rad51-NTD exhibit similar autonomous PEE activity. Mec1/Tel1-dependent phosphorylation of Rad51-NTD antagonizes the proteasomal degradation pathway, increasing the half-life of Rad51 from ∼30 min to ≥180 min (1). (page 1, lines 11-14)

Second, in our preprint manuscript, we have already shown that phosphorylation-defective Rad53-SCD1 (Rad51-SCD1-5STA) also exhibits autonomous PEE activity similar to that of wild-type Rad53-SCD (Figure 2D, Figure 4A and Figure 4C). We have highlighted this point in our revised manuscript (page 9, lines 19-21).

Third, as revealed by the results of Figure 4, it is the percentages, and not the numbers, of S/T/Q/N residues that are correlated with the PEE activities of Q-rich motifs.

The authors also do not discuss if the N-end rule for protein stability applies to the lacZ reporter or the fusion proteins.

The autonomous PEE function of S/T/Q-rich NTDs is unlikely to be relevant to the N-end rule. The N-end rule links the in vivo half-life of a protein to the identity of its N-terminal residues. In S. cerevisiae, the N-end rule operates as part of the ubiquitin system and comprises two pathways. First, the Arg/N-end rule pathway, involving a single N-terminal amidohydrolase Nta1, mediates deamidation of N-terminal asparagine (N) and glutamine (Q) into aspartate (D) and glutamate (E), which in turn are arginylated by a single Ate1 R-transferase, generating the Arg/N degron. N-terminal R and other primary degrons are recognized by a single N-recognin Ubr1 in concert with ubiquitin-conjugating Ubc2/Rad6. Ubr1 can also recognize several other N-terminal residues, including lysine (K), histidine (H), phenylalanine (F), tryptophan (W), leucine (L) and isoleucine (I) (68-70). Second, the Ac/N-end rule pathway targets proteins containing N-terminally acetylated (Ac) residues. Prior to acetylation, the first amino acid methionine (M) is catalytically removed by Met-aminopeptidases (MetAPs), unless a residue at position 2 is non-permissive (too large) for MetAPs. If a retained N-terminal M or otherwise a valine (V), cysteine (C), alanine (A), serine (S) or threonine (T) residue is followed by residues that allow N-terminal acetylation, the proteins containing these AcN degrons are targeted for ubiquitylation and proteasome-mediated degradation by the Doa10 E3 ligase (71).

The PEE activities of these S/T/Q-rich domains are unlikely to arise from counteracting the N-end rule for two reasons. First, the first two amino acid residues of Rad51-NTD, Hop1-SCD, Rad53-SCD1, Sup35-PND, Rad51-ΔN, and LacZ-NVH are MS, ME, ME, MS, ME, and MI, respectively, where M is methionine, S is serine, E is glutamic acid and I is isoleucine. Second, Sml1-NTD behaves similarly to these N-terminal fusion tags, despite its methionine and glutamine (MQ) amino acid signature at the N-terminus. (Page 12, line 3 to page 13, line 2)

The most interesting part of the paper is an exploration of S/T/Q/N-rich regions and other repetitive AA runs in 27 proteomes, particularly ciliates. However, this analysis is missing a critical control that makes it nearly impossible to evaluate the importance of the findings. The authors find the abundance of different amino acid runs in various proteomes. They also report the background abundance of each amino acid. They do not use this background abundance to normalize the runs of amino acids to create a null expectation from each proteome. For example, it has been clear for some time (Ruff, 2017; Ruff et al., 2016) that Drosophila contains a very high background of Q's in the proteome and it is necessary to control for this background abundance when finding runs of Q's.

We apologize for not explaining sufficiently well the topic eliciting this reviewer’s concern in our preprint manuscript. In the second paragraph of page 14, we cite six references to highlight that SCDs are overrepresented in yeast and human proteins involved in several biological processes (5, 43) and that polyX prevalence differs among species (79-82).

We will cite a reference by Kiersten M. Ruff in our revised manuscript (38).

K. M. Ruff, J. B. Warner, A. Posey and P. S. Tan (2017) Polyglutamine length dependent structural properties and phase behavior of huntingtin exon1. Biophysical Journal 112, 511a.

The authors could easily address this problem with the data and analysis they have already collected. However, at this time, without this normalization, I am hesitant to trust the lists of proteins with long runs of amino acid and the ensuing GO enrichment analysis. Ruff KM. 2017. Washington University in St.

Ruff KM, Holehouse AS, Richardson MGO, Pappu RV. 2016. Proteomic and Biophysical Analysis of Polar Tracts. Biophys J 110:556a.

We thank Reviewer #1 for this helpful suggestion and now address this issue by means of a different approach described below.

Based on a previous study (43), we applied seven different thresholds to seek both short and long, as well as pure and impure, polyX strings in 20 different representative near-complete proteomes, including 4X (4/4), 5X (4/5-5/5), 6X (4/6-6/6), 7X (4/7-7/7), 8-10X (≥50%X), 11-10X (≥50%X) and ≥21X (≥50%X).

To normalize the runs of amino acids and create a null expectation from each proteome, we determined the ratios of the overall number of X residues for each of the seven polyX motifs relative to those in the entire proteome of each species, respectively. The results of four different polyX motifs are shown in our revised manuscript, i.e., polyQ (Figure 7), polyN (Figure 8), polyS (Figure 9) and polyT (Figure 10). Thus, polyX prevalence differs among species and the overall X contents of polyX motifs often but not always correlate with the X usage frequency in entire proteomes (43).

Most importantly, our results reveal that, compared to Stentor coeruleus or several non-ciliate eukaryotic organisms (e.g., Plasmodium falciparum, Caenorhabditis elegans, Danio rerio, Mus musculus and Homo sapiens), the five ciliates with reassigned TAAQ and TAGQ codons not only have higher Q usage frequencies, but also more polyQ motifs in their proteomes (Figure 7). In contrast, polyQ motifs prevail in Candida albicans, Candida tropicalis, Dictyostelium discoideum, Chlamydomonas reinhardtii, Drosophila melanogaster and Aedes aegypti, though the Q usage frequencies in their entire proteomes are not significantly higher than those of other eukaryotes (Figure 1). Due to their higher N usage frequencies, Dictyostelium discoideum, Plasmodium falciparum and Pseudocohnilembus persalinus have more polyN motifs than the other 23 eukaryotes we examined here (Figure 8). Generally speaking, all 26 eukaryotes we assessed have similar S usage frequencies and percentages of S contents in polyS motifs (Figure 9). Among these 26 eukaryotes, Dictyostelium discoideum possesses many more polyT motifs, though its T usage frequency is similar to that of the other 25 eukaryotes (Figure 10).

In conclusion, these new normalized results confirm that the reassignment of stop codons to Q indeed results in both higher Q usage frequencies and more polyQ motifs in ciliates.

Reviewer #2 (Public Review):

Summary:

This study seeks to understand the connection between protein sequence and function in disordered regions enriched in polar amino acids (specifically Q, N, S and T). While the authors suggest that specific motifs facilitate protein-enhancing activities, their findings are correlative, and the evidence is incomplete. Similarly, the authors propose that the re-assignment of stop codons to glutamine-encoding codons underlies the greater user of glutamine in a subset of ciliates, but again, the conclusions here are, at best, correlative. The authors perform extensive bioinformatic analysis, with detailed (albeit somewhat ad hoc) discussion on a number of proteins. Overall, the results presented here are interesting, but are unable to exclude competing hypotheses.

Strengths:

Following up on previous work, the authors wish to uncover a mechanism associated with poly-Q and SCD motifs explaining proposed protein expression-enhancing activities. They note that these motifs often occur IDRs and hypothesize that structural plasticity could be capitalized upon as a mechanism of diversification in evolution. To investigate this further, they employ bioinformatics to investigate the sequence features of proteomes of 27 eukaryotes. They deepen their sequence space exploration uncovering sub-phylum-specific features associated with species in which a stop-codon substitution has occurred. The authors propose this stop-codon substitution underlies an expansion of ploy-Q repeats and increased glutamine distribution.

Weaknesses:

The preprint provides extensive, detailed, and entirely unnecessary background information throughout, hampering reading and making it difficult to understand the ideas being proposed.

The introduction provides a large amount of detailed background that appears entirely irrelevant for the paper. Many places detailed discussions on specific proteins that are likely of interest to the authors occur, yet without context, this does not enhance the paper for the reader.

The paper uses many unnecessary, new, or redefined acronyms which makes reading difficult. As examples:

1. Prion forming domains (PFDs). Do the authors mean prion-like domains (PLDs), an established term with an empirical definition from the PLAAC algorithm? If yes, they should say this. If not, they must define what a prion-forming domain is formally.

The N-terminal domain (1-123 amino acids) of S. cerevisiae Sup35 was already referred to as a “prion forming domain (PFD)” in 2006 (48). Since then, PFD has also been employed as an acronym in other yeast prion papers (Cox, B.S. et al. 2007; Toombs, T. et al. 2011).

B. S. Cox, L. Byrne, M. F., Tuite, Protein Stability. Prion 1, 170-178 (2007). J. A. Toombs, N. M. Liss, K. R. Cobble, Z. Ben-Musa, E. D. Ross, [PSI+] maintenance is dependent on the composition, not primary sequence, of the oligopeptide repeat domain. PLoS One 6, e21953 (2011).

1. SCD is already an acronym in the IDP field (meaning sequence charge decoration) - the authors should avoid this as their chosen acronym for Serine(S) / threonine (T)-glutamine (Q) cluster domains. Moreover, do we really need another acronym here (we do not).

SCD was first used in 2005 as an acronym for the Serine (S)/threonine (T)-glutamine (Q) cluster domain in the DNA damage checkpoint field (4). Almost a decade later, SCD became an acronym for “sequence charge decoration” (Sawle, L. et al. 2015; Firman, T. et al. 2018).

L. Sawle and K, Ghosh, A theoretical method to compute sequence dependent configurational properties in charged polymers and proteins. J. Chem Phys. 143, 085101(2015).

T. Firman and Ghosh, K. Sequence charge decoration dictates coil-globule transition in intrinsically disordered proteins. J. Chem Phys. 148, 123305 (2018).

1. Protein expression-enhancing (PEE) - just say expression-enhancing, there is no need for an acronym here.

Thank you. Since we have shown that the addition of Q-rich motifs to LacZ affects protein expression rather than transcription, we think it is better to use the “PEE” acronym.

The results suggest autonomous protein expression-enhancing activities of regions of multiple proteins containing Q-rich and SCD motifs. Their definition of expression-enhancing activities is vague and the evidence they provide to support the claim is weak. While their previous work may support their claim with more evidence, it should be explained in more detail. The assay they choose is a fusion reporter measuring beta-galactosidase activity and tracking expression levels. Given the presented data they have shown that they can drive the expression of their reporters and that beta gal remains active, in addition to the increase in expression of fusion reporter during the stress response. They have not detailed what their control and mock treatment is, which makes complete understanding of their experimental approach difficult. Furthermore, their nuclear localization signal on the tag could be influencing the degradation kinetics or sequestering the reporter, leading to its accumulation and the appearance of enhanced expression. Their evidence refuting ubiquitin-mediated degradation does not have a convincing control.

Although this reviewer’s concern regarding our use of a nuclear localization signal on the tag is understandable, we are confident that this signal does not bias our findings for two reasons. First, the negative control LacZ-NV also possesses the same nuclear localization signal (Figure 1A, lane 2). Second, another fusion target, Rad51-ΔN, does not harbor the NVH tag (Figure 1D, lanes 3-4). Compared to wild-type Rad51, Rad51-ΔN is highly labile. In our previous study, removal of the NTD from Rad51 reduced by ~97% the protein levels of corresponding Rad51-ΔN proteins relative to wild-type (1).

Based on the experimental results, the authors then go on to perform bioinformatic analysis of SCD proteins and polyX proteins. Unfortunately, there is no clear hypothesis for what is being tested; there is a vague sense of investigating polyX/SCD regions, but I did not find the connection between the first and section compelling (especially given polar-rich regions have been shown to engage in many different functions). As such, this bioinformatic analysis largely presents as many lists of percentages without any meaningful interpretation. The bioinformatics analysis lacks any kind of rigorous statistical tests, making it difficult to evaluate the conclusions drawn. The methods section is severely lacking. Specifically, many of the methods require the reader to read many other papers. While referencing prior work is of course, important, the authors should ensure the methods in this paper provide the details needed to allow a reader to evaluate the work being presented. As it stands, this is not the case.

Thank you. As described in detail below, we have now performed rigorous statistical testing using the GofuncR package (Figure 11, Figure 12 and DS7-DS32).

Overall, my major concern with this work is that the authors make two central claims in this paper (as per the Discussion). The authors claim that Q-rich motifs enhance protein expression. The implication here is that Q-rich motif IDRs are special, but this is not tested. As such, they cannot exclude the competing hypothesis ("N-terminal disordered regions enhance expression").

In fact, “N-terminal disordered regions enhance expression” exactly summarizes our hypothesis.

On pages 12-13 and Figure 4 of our preprint manuscript, we explained our hypothesis in the paragraph entitled “The relationship between PEE function, amino acid contents, and structural flexibility”.

The authors also do not explore the possibility that this effect is in part/entirely driven by mRNA-level effects (see Verma Na Comms 2019).

As pointed out by the first reviewer, we present evidence that the increase in protein abundance and enzymatic activity is not due to changes in plasmid copy number or mRNA abundance (Figure 2), and that this phenomenon is not affected in translational quality control mutants (Figure 3).

As such, while these observations are interesting, they feel preliminary and, in my opinion, cannot be used to draw hard conclusions on how N-terminal IDR sequence features influence protein expression. This does not mean the authors are necessarily wrong, but from the data presented here, I do not believe strong conclusions can be drawn. That re-assignment of stop codons to Q increases proteome-wide Q usage. I was unable to understand what result led the authors to this conclusion.

My reading of the results is that a subset of ciliates has re-assigned UAA and UAG from the stop codon to Q. Those ciliates have more polyQ-containing proteins. However, they also have more polyN-containing proteins and proteins enriched in S/T-Q clusters. Surely if this were a stop-codon-dependent effect, we'd ONLY see an enhancement in Q-richness, not a corresponding enhancement in all polar-rich IDR frequencies? It seems the better working hypothesis is that free-floating climate proteomes are enriched in polar amino acids compared to sessile ciliates.

We thank this reviewer for raising this point, however her/his comments are not supported by the results in Figure 7.

Regardless, the absence of any kind of statistical analysis makes it hard to draw strong conclusions here.

We apologize for not explaining more clearly the results of Tables 5-7 in our preprint manuscript.

To address the concerns about our GO enrichment analysis by both reviewers, we have now performed rigorous statistical testing for SCD and polyQ protein overrepresentation using the GOfuncR package (https://bioconductor.org/packages/release/bioc/html/GOfuncR.html). GOfuncR is an R package program that conducts standard candidate vs. background enrichment analysis by means of the hypergeometric test. We then adjusted the raw p-values according to the Family-wise error rate (FWER). The same method had been applied to GO enrichment analysis of human genomes (89).

The results presented in Figure 11 and Figure 12 (DS7-DS32) support our hypothesis that Q-rich motifs prevail in proteins involved in specialized biological processes, including Saccharomyces cerevisiae RNA-mediated transposition, Candida albicans filamentous growth, peptidyl-glutamic acid modification in ciliates with reassigned stop codons (TAAQ and TAGQ), Tetrahymena thermophila xylan catabolism, Dictyostelium discoideum sexual reproduction, Plasmodium falciparum infection, as well as the nervous systems of Drosophila melanogaster, Mus musculus, and Homo sapiens (78). In contrast, peptidyl-glutamic acid modification and microtubule-based movement are not overrepresented with Q-rich proteins in Stentor coeruleus, a ciliate with standard stop codons.

Recommendations for the authors:

Please note that you control which revisions to undertake from the public reviews and recommendations for the authors.

Reviewer #1 (Recommendations For The Authors):

The order of paragraphs in the introduction was very difficult to follow. Each paragraph was clear and easy to understand, but the order of paragraphs did not make sense to this reader. The order of events in the abstract matches the order of events in the results section. However, the order of paragraphs in the introduction is completely different and this was very confusing. This disordered list of facts might make sense to an expert reader but makes it hard for a non-expert reader to understand.

Apologies. We endeavored to improve the flow of our revised manuscript to make it more readable.

The section beginning on pg 12 focused on figures 4 and 5 was very interesting and highly promising. However, it was initially hard for me to tell from the main text what the experiment was. Please add to the text an explanation of the experiment, because it is hard to figure out what was going on from the figures alone. Figure 4 is fantastic, but would be improved by adding error bars and scaling the x-axis to be the same in panels B,C,D.

Thank you for this recommendation. We have now scaled both the x-axis and y-axis equivalently in panels B, C and D of Figure 4. Error bars are too small to be included.

It is hard to tell if the key variable is the number of S/T/Q/N residues or the number of phosphosites. I think a good control would be to add a regression against the number of putative phosphosites. The sequences are well designed. I loved this part but as a reader, I need more interpretation about why it matters and how it explains the PEE.

As described above, we have shown that the number of phosphorylation sites in the Q-rich motifs is not relevant to their autonomous protein expression-enhancing (PEE) activities.

I believe that the prevalence of polyX runs is not meaningful without normalizing for the background abundance of each amino acid. The proteome-wide abundance and the assumption that amino acids occur independently can be used to form a baseline expectation for which runs are longer than expected by chance. I think Figures 6 and 7 should go into the supplement and be replaced in the main text with a figure where Figure 6 is normalized by Figure 7. For example in P. falciparum, there are many N-runs (Figure 6), but the proteome has the highest fraction of N’s (Figure 7).

Thank you for these suggestions. The three figures in our preprint manuscript (Figures 6-8) have been moved into the supplementary information (Figures S1-S3). For normalization, we have provided four new figures (Figures 7-10) in our revised manuscript.

The analysis of ciliate proteomes was fascinating. I am particularly interested in the GO enrichment for “peptidyl-glutamic acid modification” (pg 20) because these enzymes might be modifying some of Q’s in the Q-runs. I might be wrong about this idea or confused about the chemistry. Do these ciliates live in Q-rich environments? Or nitrogen rich environments?

Polymeric modifications (polymodifications) are a hallmark of C-terminal tubulin tails, whereas secondary peptide chains of glutamic acids (polyglutamylation) and glycines (polyglycylation) are catalyzed from the γ-carboxyl group of primary chain glutamic acids. It is not clear if these enzymes can modify some of the Q’s in the Q-runs.

To our knowledge, ciliates are abundant in almost every liquid water environment, i.e., oceans/seas, marine sediments, lakes, ponds, and rivers, and even soils.

I think you should include more discussion about how the codons that code for Q’s are prone to slippage during DNA replication, and thus many Q-runs are unstable and expand (e.g. Huntington’s Disease). The end of pg 24 or pg 25 would be good places.

We thank the reviewer for these comments.

PolyQ motifs have a particular length-dependent codon usage that relates to strand slippage in CAG/CTG trinucleotide repeat regions during DNA replication. In most organisms having standard genetic codons, Q is encoded by CAGQ and CAAQ. Here, we have determined and compared proteome-wide Q contents, as well as the CAGQ usage frequencies (i.e., the ratio between CAGQ and the sum of CAGQ, CAGQ, TAAQ, and TAGQ).

Our results reveal that the likelihood of forming long CAG/CTG trinucleotide repeats are higher in five eukaryotes due to their higher CAGQ usage frequencies, including Drosophila melanogaster (86.6% Q), Danio rerio (74.0% Q), Mus musculus (74.0% Q), Homo sapiens (73.5% Q), and Chlamydomonas reinhardtii (87.3% Q) (orange background, Table 2). In contrast, another five eukaryotes that possess high numbers of polyQ motifs (i.e., Dictyostelium discoideum, Candida albicans, Candida tropicalis, Plasmodium falciparum and Stentor coeruleus) (Figure 1) utilize more CAAQ (96.2%, 84.6%, 84.5%, 86.7% and 75.7%) than CAAQ (3.8%, 15.4%, 15.5%, 13.3% and 24.3%), respectively, to avoid the formation of long CAG/CTG trinucleotide repeats (green background, Table 2). Similarly, all five ciliates with reassigned stop codons (TAAQ and TAGQ) have low CAGQ usage frequencies (i.e., from 3.8% Q in Pseudocohnilembus persalinus to 12.6% Q in Oxytricha trifallax) (red font, Table 2). Accordingly, the CAG-slippage mechanism might operate more frequently in Chlamydomonas reinhardtii, Drosophila melanogaster, Danio rerio, Mus musculus and Homo sapiens than in Dictyostelium discoideum, Candida albicans, Candida tropicalis, Plasmodium falciparum, Stentor coeruleus and the five ciliates with reassigned stop codons (TAAQ and TAGQ).

Pg 7, paragraph 2 has no direction. Please add the conclusion of the paragraph to the first sentence.

This paragraph has been moved to the “Introduction” section” of the revised manuscript.

Pg 8, I suggest only mentioning the PFDs used in the experiments. The rest are distracting.

We have addressed this concern above.

Pg 12. Please revise the "The relationship...." text to explain the experiment.

We apologize for not explaining this topic sufficiently well in our preprint manuscript.

SCDs are often structurally flexible sequences (4) or even IDRs. Using IUPred2A (https://iupred2a.elte.hu/plot_new), a web-server for identifying disordered protein regions (88), we found that Rad51-NTD (1-66 a.a.) (1), Rad53-SCD1 (1-29 a.a.) and Sup35-NPD (1-39 a.a.) are highly structurally flexible. Since a high content of serine (S), threonine (T), glutamine (Q), asparanine (N) is a common feature of IDRs (17-20), we applied alanine scanning mutagenesis approach to reduce the percentages of S, T, Q or N in Rad51-NTD, Rad53-SCD1 or Sup35-NPD, respectively. As shown in Figure 4 and Figure 5, there is a very strong positive relationship between STQ and STQN amino acid percentages and β-galactosidase activities. (Page 13, lines 5-10)

Pg 13, first full paragraph, "Futionally, IDRs..." I think this paragraph belongs in the Discussion.

This paragraph is now in the “Introduction” section (Page 5, Lines 11-15).

Pg. 15, I think the order of paragraphs should be swapped.

These paragraphs have been removed or rewritten in the “Introduction section” of our revised manuscript.

Pg 17 (and other parts) I found the lists of numbers and percentages hard to read and I think you should refer readers to the tables.

Thank you. In the revised manuscript, we have avoided using lists of numbers and percentages, unless we feel they are absolutely essential.

Pg. 19 please add more interpretation to the last paragraph. It is very cool but I need help understanding the result. Are these proteins diverging rapidly? Perhaps this is a place to include the idea of codon slippage during DNA replication.

Thank you. The new results in Table 2 indicate that the CAG-slippage mechanism is unlikely to operate in ciliates with reassigned stop codons (TAAQ and TAGQ).

Pg 24. "Based on our findings from this study, we suggest that Q-rich motifs are useful toolkits for generating novel diversity during protein evolution, including by enabling greater protein expression, protein-protein interactions, posttranslational modifications, increased solubility, and tunable stability, among other important traits." This idea needs to be cited. Keith Dunker has written extensively about this idea as have others. Perhaps also discuss why Poly Q rich regions are different from other IDRs and different from other IDRs that phase-separate.

Agreed, we have cited two of Keith Dunker’s papers in our revised manuscript (73, 74).

Minor notes:

Please define Borg genomes (pg 25).

Borgs are long extrachromosomal DNA sequences in methane-oxidizing Methanoperedens archaea, which display the potential to augment methane oxidation (101). They are now described in our revised manuscript. (Page 15, lines 12-14)

Reviewer #2 (Recommendations For The Authors):

The authors dance around disorder but never really quantify or show data. This seems like a strange blindspot.

We apologize for not explaining this topic sufficiently well in our preprint manuscript. We have endeavored to do so in our revised manuscript.

The authors claim the expression enhancement is "autonomous," but they have not ruled things out that would make it not autonomous.

Evidence of the “autonomous” nature of expression enhancement is presented in Figure 1, Figure 4, and Figure 5 of the preprint manuscript.

Recommendations for improving the writing and presentation.

The title does not recapitulate the entire body of work. The first 5 figures are not represented by the title in any way, and indeed, I have serious misgivings as to whether the conclusion stated in the title is supported by the work. I would strongly suggest the authors change the title.

Figure 2 could be supplemental.

Thank you. We think it is important to keep Figure 2 in the text.

Figures 4 and 5 are not discussed much or particularly well.

This reviewer’s opinion of Figure 4 and Figure 5 is in stark contrast to those of the first reviewer.

The introduction, while very thorough, takes away from the main findings of the paper. It is more suited to a review and not a tailored set of minimal information necessary to set up the question and findings of the paper. The question that the authors are after is also not very clear.

Thank you. The entire “Introduction” section has been extensively rewritten in the revised manuscript.

Schematics of their fusion constructs and changes to the sequence would be nice, even if supplemental.

Schematics of the fusion constructs are provided in Figure 1A.

The methods section should be substantially expanded.

The method section in the revised manuscript has been rewritten and expanded. The six Javascript programs used in this work are listed in Table S4.

The text is not always suited to the general audience and readership of eLife.

We have now rewritten parts of our manuscript to make it more accessible to the broad readership of eLife.

In some cases, section headers really don't match what is presented, or there is no evidence to back the claim.

The section headers in the revised manuscript have been corrected.

A lot of the listed results in the back half of the paper could be a supplemental table, listing %s in a paragraph (several of them in a row) is never nice

Acknowledged. In the revised manuscript, we have removed almost all sentences listing %s.

Minor corrections to the text and figures.

There is a reference to table 1 multiple times, and it seems that there is a missing table. The current table 1 does not seem to be the same table referred to in some places throughout the text.

Apologies for this mistake, which we have now corrected in our revised manuscript.

In some places its not clear where new work is and where previous work is mentioned. It would help if the authors clearly stated "In previous work...."

Acknowledged. We have corrected this oversight in our revised manuscript.

Not all strains are listed in the strain table (KO's in figure 3 are not included)

Apologies, we have now corrected Table S2, as suggested by this reviewer.

Version 2

Abstract

Abstract

Serine(S)/threonine(T)-glutamine(Q) cluster domains (SCDs), polyglutamine (polyQ) tracts and polyglutamine/asparagine (polyQ/N) tracts are Q-rich motifs found in many proteins. SCDs often are intrinsically disordered regions that mediate protein phosphorylation and protein-protein interactions. PolyQ and polyQ/N tracts are structurally flexible sequences that trigger protein aggregation. We report that due to their high percentages of STQ or STQN amino acid content, four SCDs and three prion-causing Q/N-rich motifs of yeast proteins possess autonomous protein expression-enhancing activities. Since these Q-rich motifs can endow proteins with structural and functional plasticity, we suggest that they represent useful toolkits for evolutionary novelty. Comparative Gene Ontology (GO) analyses of the near-complete proteomes of 27 representative model eukaryotes reveal that Q-rich motifs prevail in proteins involved in specialized biological processes, including Saccharomyces cerevisiae RNA-mediated transposition and pseudohyphal growth, Candida albicans filamentous growth, ciliate peptidyl-glutamic acid modification and microtubule-based movement, Tetrahymena thermophila xylan catabolism and meiosis, Dictyostelium discoideum development and sexual cycles, Plasmodium falciparum infection, and the nervous systems of Drosophila melanogaster, Mus musculus and Homo sapiens. We also show that Q-rich-motif proteins are expanded massively in ten ciliates with reassigned TAAQ and TAGQ codons. Notably, the usage frequency of CAGQ is much lower in ciliates with reassigned TAAQ and TAGQ codons than in organisms with expanded and unstable Q runs (e.g., D. melanogaster and H. sapiens), indicating that the use of noncanonical stop codons in ciliates may have coevolved with codon usage biases to avoid triplet repeat disorders mediated by CAG/GTC replication slippage.

Introduction

We reported previously that the NH2-terminal domain (NTD; residues 1–66) of budding yeast Saccharomyces cerevisiae Rad51 protein contains three SQ motifs (S2Q, S12Qand S30Q) (1). The S/T-Q motifs, comprising S or T followed by Q, are the target sites of DNA damage sensor protein kinases, i.e., ATM (ataxia-telangiectasia mutated), ATR (RAD3-related) (2, 3) and DNA-dependent protein kinase (DNA-PK) (4, 5). Mec1 (Mitotic Entry Checkpoint 1) and Tel1 (TELomere maintenance 1) are the budding yeast homologs of mammalian ATR and ATM, respectively. Budding yeast lacks a DNA-PK homolog (2, 3, 6). This clustering of three SQ motifs within a stretch of 31 amino acids in Rad51-NTD fulfills the criteria to define an S/T-Q cluster domain (SCD) (4, 5). The three SQ motifs of Rad51-NTD are phosphorylated in a Mec1-and Tel1-dependent manner during vegetative growth and meiosis. Mec1/Tel1-dependent NTD phosphorylation antagonizes Rad51 degradation via the proteasomal pathway, increasing the half-life of Rad51 from ∼30 min to ≥ 180 min (1), supporting the notion that Mec1 and Tel1 exhibit an essential function in regulating protein homeostasis (proteostasis) in S. cerevisiae (7, 8).

A unifying definition of an SCD is having ≥3 S/T-Q sites within a stretch of 50-100 amino acids (4, 5). One of the best-understood mechanisms of SCD phosphorylation involves the association of SCDs with their binding partners containing a forkhead-associated (FHA) domain. For example, Mec1/Tel1-dependent phosphorylation of Rad53-SCD1 (residues 1-29) and Hop1-SCD (residues 258-324) specifically recruits and activates their downstream DNA damage checkpoint kinases Dun1 and Mek1, respectively (9-11). Dun1 phosphorylates three serine residues (S56, S58 and S60) of the ribonucleotide reductase inhibitor Sml1, subsequently promoting Sml1 ubiquitination by the E2 ubiquitin-conjugating enzyme Rad6 and the E3 ubiquitin ligase Ubr2, as well as promoting Sml1 degradation via the 26S proteasome (12–14). Mek1 phosphorylates two Rad51 accessory factors, Rad54 and Hed1 (a meiosis-specific inhibitor of Rad51), suppressing Rad51’s strand-exchange activity and preventing Rad51-mediated DSB repair, respectively (15, 16).

There are many other SCD-containing proteins that are neither targets of ATM/Tel1 or ATR/Mec1 nor functionally linked to DNA Damage response or DNA repair (4, 5), indicating that SCDs might possess previously uncharacterized biochemical properties or physiological functions. Interestingly, due to their high percentages of STQ amino acid content, SCDs often are intrinsically disordered regions (IDRs) in their native states rather than adopting stable secondary and/or tertiary structures (4). A common feature of IDRs is their high content of serine (S), threonine (T), glutamine (Q), asparanine (N), proline (P), glycine (G) or charged amino acids [arginine (R), lysine (K) and histidine (H)] (17–20). Functionally, IDRs are key components of subcellular machineries and signaling pathways because they have the potential to associate with many partners due to their multiple possible metastable conformations. Many IDRs are regulated by alternative splicing and post-translational modifications. Some IDRs are involved in the formation of various membraneless organelles via intracellular liquid-liquid phase separation (21, 22). Highly-charged IDRs can act as entropic bristles that, when translationally fused to their partner proteins, only enhance the water solubility but not the steady-state levels of their partner proteins (23).

In this study, we first report that seven Q-rich motifs of S. cerevisiae proteins, including Rad51-NTD (1), have high STQN or STQ amino acid contents and exhibit autonomous expression-enhancing activity for high-level production of native protein and when fused to exogenous target proteins, e.g., β-galactosidase (LacZ), in vivo. We also reveal structural and genetic requirements for the “nanny” function of these Q-rich motifs in regulating protein homeostasis, leading to the hypothesis that Q-rich motifs are useful toolkits for structural and functional plasticity, as well as evolutionary novelty. Next, we performed Gene Ontology (GO) enrichment analyses on all proteins having Q-rich motifs (i.e., SCDs, polyQ and polyQ/N), as well as those with the homorepeat (polyX) motifs of other amino acid residues, in 20 non-ciliate and 17 ciliate species. Notably, relative to most other eukaryotes, many ciliates reassign their standard stop codons into amino acids (Table 1). For example, several ciliates possess two noncanonical nuclear genetic codes (UAAQ and UAGQ), in which the UAA and UAG stop codons have been reassigned to glutamine (Q) so that UGA is the sole functional stop codon, including Tetrahymena thermophila, Paramecium tetraurelia, Paramecium bursaria, Oxytricha trifallax, Stylonychia lemnae, Pseudocohnilembus persalinus, Aristerostoma sp., Favella ehrenbergii, Pseudokeronopsis spp., Strombidium inclinatum, and Uronema spp. Both the UAA and UAG stop codons are reassigned to tyrosine (Y) in Favella ehrenbergii, whereas the UGA stop codon is translated to tryptophan (W) or cysteine (C) in Blepharisma japonicum and Euplotes focardii, respectively. In contrast, Stentor coeruleus, Climacostomum virens, Litonotus pictus and Protocruzia adherens utilize the universal set of genetic codons. Condylostoma magnum and Parduzcia sp. have no dedicated genetic codes. Their UAA, UAG and UGA codons can be stop codons or translated to Q, C and W, respectively. Translation termination at the mRNA 3′ end occurs in a context-dependent manner to distinguish stop from sense (24–36). Although it has been reported previously that Q is used more frequently in Tetrahymena thermophila and Paramecium tetraurelia than in other species (29, 30), many important questions regarding stop codon reassignment in ciliates remain unresolved. For instance, fundamentally, it is unclear if Q, Y, W and C are used more frequently in other ciliates in which stop codons are reassigned. Moreover, whether there are common or specific structural motif(s) in proteins arising from stop codon reassignment is not clear. Furthermore, what are the structural and functional impacts of such genome-wide alterations? Finally, the codons that code for the polyQ motifs are prone to a CAG/GTC-slippage mechanism during DNA replication, and thus many Q-runs are unstable and expanded in some eukaryotic organisms, leading to polyQ-associated diseases (e.g., Huntington’s Disease) (37–39). Accordingly, polyQ tracts in proteins associated with disease are more enriched in the CAG codon, becoming almost CAG exclusive (39, 40). In contrast, CAA insertions in a Drosophila model of a polyQ-associated disease revealed that even though the resulting polyQ tract is of the same length as the disease-associated tract, the proteins display reduced toxicity (41). In this study, we determine and compare the usage frequency of TAAQ, TAGQ, CAAQ, and CAGQ in ciliates and non-ciliate eukaryotes.

Results

SCDs provide versatile functionalities in proteins

We have shown previously in S. cerevisiae that Rad51-NTD autonomously promotes high-level production of native Rad51 and its COOH-terminal fusion protein LacZ (β-galactosidase) in vivo (1). To do so, in brief, we expressed Rad51-NTD-LacZ-NVH fusion proteins using a CEN-ARS plasmid (low-copy number) under the control of the native RAD51 gene promoter (PRAD51) (Table 1). The NVH tag contains an SV40 nuclear localization signal (NLS) peptide preceding a V5 epitope tag and a hexahistidine (His6) affinity tag (1). We confirmed that the N-terminal addition of Rad51-NTD to LacZ-NVH increased both steady-state levels of LacZ-NVH fusion proteins (Figure 1A) and β-galactosidase activities in vivo (Figure 1B). Here, we further report that yeast Rad53-SCD1, Hop1-SCD, Sml1-NTD1–50 (residues 1–50) and Sml1-NTD1–27 (residues 1–27) also exhibit protein expression-enhancing (PEE) activities (Figure 1A and 1B, Table 1). The Sml1 protein in the SK1 strain harbors three S/T-Q motifs (S4Q, S14Q and T47Q), whereas that in the S288c strain only has one SQ motif (S4Q, C14Q and T47M).

The Q-rich domains of seven different yeast proteins possess autonomous expression-enhancing (PEE) activities.

(A-B) N-terminal fusion of Rad51-NTD/SCD, Rad53-SCD1, Hop1-SCD, Sml1-NTD, Sup35-PND, Ure2-UPD and New1-NPD promotes high-level expression of LacZ-NVH, respectively. The NVH tag contains an SV40 nuclear localization signal (NLS) peptide preceding a V5 epitope tag and a hexahistidine (His6) affinity tag (1). Western blots for visualization of LacZ-NVH fusion proteins (A) and quantitative β-galactosidase assays (B) were carried out as described previously (1). Error bars indicate standard deviation between experiments (n ≥ 3). Asterisks indicate significant differences relative to wild type (WT) in A or lacking an NTD in B, with P values calculated using a two-tailed t-test (***, P value <0.001; **, P value <0.01). (C-D) The PEE activities of S/T/Q/N-rich domains are independent of the quaternary structures of target proteins. (C) Rad53-SCD1 can be used as an N-terminal fusion tag to enhance production of four different target proteins: LacZ-NVH, GST-NVH, GSTnd-NVH and GFP-NVH. (D) Visualization of native Rad51 (NTD-Rad51-ΔN), Rad51-ΔN, and the Rad51-ΔN fusion proteins by immunoblotting. Hsp104 was used as a loading control. Size in kilodaltons of standard protein markers is labeled to the left of the blots. The black arrowhead indicates the protein band of Rad51-ΔN. (E) MMS sensitivity. Spot assay showing five-fold serial dilutions of indicated strains grown on YPD plates with or without MMS at the indicated concentrations (w/v).

The Q-rich motifs of three yeast prion-causing proteins also exhibit PEE activities

Since Sml1-NTD1–27 in the SK1 strain only harbors two S/T-Q motifs (S4Q and S14Q), the number of S/T-Q motifs alone could not account for PEE activity. Notably, Rad51-NTD, Rad53-SCD1, Hop1-SCD, Sml1-NTD1-27 and Sml1-NTD1-50 all represent Q- or Q/N-rich motifs. Rad51-NTD contains 9 serines (S), 2 threonines (T), 9 glutamines (Q) and 4 asparagines (N). Rad53-SCD1 has 2S, 4T, 7Q and 1N. Hop1-SCD has 6S, 6T, 8Q and 9N. Sml1-NTD1–27 and Sml1-NTD1–50 in SK1 possess 3S and 5S, 2T and 3T, 6Q and 7Q, as well as 2N and 3N, respectively.

Accordingly, we investigated if other Q- or Q/N-rich motifs in yeast can also promote protein expression in vivo. PolyQ and polyQ/N tracts are the most common homorepeats acting as structurally flexible motifs for protein aggregation or protein–protein interactions in many eukaryotes (42, 43). PolyN is not as structurally flexible as polyQ due to a stronger propensity for β-turn formation in polyN than in polyQ (44). In so-called polyQ-associated diseases, long Q-, Q/N- or even N-rich motifs result in an excess of interactions, resulting in dysfunctional or pathogenic protein aggregates (45). Many prion-causing proteins contain Q/N-rich prion-forming domains (PFDs). In S. cerevisiae, the best-characterized prion-causing proteins are Sup35 (or translation terminator eRF35), New1 ([NU+] prion formation protein 1), Ure2 (uridosuccinate transport 2), Rnq1 (rich in N and Q 1) and Swi1 (switching deficient 1) (46, 47). We found that the Q/N-rich NTDs of Sup35, Ure2 and New1 also display PEE activities, i.e., the prion nucleation domain (PND; residues 1-39) of Sup35 (48), the Ure2 prion domain (UPD) (residues 1-91) (49, 50), and the New1 prion domain (NPD; residue 1-146) (51) (Figure 1A and 1B, Table S1). Sup35-PND containing 3S, 12Q, 18N and an S17Q motif exerts critical functions in promoting [PSI+] prion nucleation (52). The UPD of the Ure2 nitrogen catabolite repression transcriptional regulator is the basis of the prion [URE3+] (49, 50). The UPD is critical for Ure2’s function in vivo because its removal in the corresponding Ure2-ΔUPD mutants elicits reduced protein stability and steady-state protein levels (but not transcript levels) (51). Ure2-UPD contains 10S, 5T, 10Q and 33N, adopting a completely disordered structure (53). New1 is a non-essential ATP-binding cassette type F protein that fine-tunes the efficiency of translation termination or ribosome recycling (54). The NPD of New1 supports [NU+] and is susceptible to [PSI+] prion induction (55, 56). New1-NPD contains 19S, 8T, 14Q, 28N and an S145Q motif. Here, we applied the LacZ-NVH fusion protein approach to show that N-terminal fusion of Sup35-PND, Ure2-UPD or New1-NPD to LacZ-NVH all increased steady-state protein levels (Figure 1A) and β-galactosidase activities in vivo (Figure 1B).

The PEE function is not affected by the quaternary structures of target proteins

We found that N-terminal fusion of Rad53-SCD1 to four different NVH-tagged target proteins (Figure 1C) or Rad51-ΔN (Figure 1D) all resulted in higher protein production in vivo. LacZ is a tetrameric protein, glutathione S-transferase (GST) is dimeric, and non-dimerizing GST (GSTnd) and GFP are monomeric proteins. As reported recently (1), removal of the NTD from Rad51 reduced by ∼97% the levels of corresponding Rad51-ΔN proteins relative to wild type (WT) (Figure 1D), leading to lower resistance to the DNA damage agent methyl methanesulfonate (MMS) (Figure 1E). Interestingly, the autonomous PEE function of Rad51-NTD could be fully rescued in rad51-ΔN (Table S2) by N-terminal fusion of Rad53-SCD1, Rad53-SCD1-5STA (all five S/T-Q motifs changed to AQs) or Sup35-PND, respectively. Rad53-SCD1-5STA is a mutant protein defective in Mec1- and Tel1-mediated phosphorylation. Compared to wild-type yeast cells, the three corresponding yeast mutants (rad53-SCD1-rad51-ΔN, rad53-SCD1-5STA-rad51-ΔN and sup35-PND-rad51-ΔN) not only produced similar steady-state levels of Rad51-ΔN fusion proteins (Figure 1D), but they also exhibited high MMS resistance (Figure 1E).

During homology-directed repair of DNA double-strand breaks (DSBs), Rad51 polymerizes into helical filaments on DSB-associated single-stranded DNA (ssDNA) and then promotes homologous search and strand exchange of the ssDNA-protein filament with a second double-stranded DNA (dsDNA). We inferred that the catalytic activity of Rad51-ΔN during DSB repair is likely similar to that of wild-type Rad51 because the weak MMS-resistant phenotype of rad51-ΔN is mainly due to very low steady-state levels of Rad51-ΔN (Figure 1D).

In conclusion, our results indicate that the quaternary structures of the target proteins (i.e., GFP, GSTnd, GST, LacZ and Rad51-ΔN) are irrelevant to the autonomous PEE activity. We assert that our use of a nuclear localization signal on the C-terminal VHN tag was unlikely to influence protein degradation kinetics or to sequester the reporter, leading to their accumulation and the appearance of enhanced expression for two reasons. First, the negative control LacZ-NV also possesses the same nuclear localization signal (Figure 1A, lane 2). Second, as an endogenous fusion target, Rad51-ΔN does not harbor the NVH tag (Figure 1D, lanes 3-4). Compared to wild-type Rad51, Rad51-ΔN is highly labile. In our previous study, removal of the NTD from Rad51 reduced by ∼97% the protein levels of corresponding Rad51-ΔN proteins relative to wild-type (1).

The autonomous PEE function is not likely controlled by plasmid copy number or its transcription

The PEE function is unlikely to operate at the transcriptional level, as revealed by genomic and reverse-transcription quantitative polymerase chain reaction analyses (i.e., g-qPCR and RT-qPCR, respectively) (Figure 2, Table S3 and DS1). We found that the addition of WT and mutant Rad51-NTD to LacZ-NVH not only did not affect the average copy number of the corresponding CEN-ARS plasmids in exponentially growing S. cerevisiae cells (Figure 2A), but also even reduced the steady-state transcript levels of the corresponding LacZ-NVH fusion protein genes (Figure 2B). Therefore, the addition of Rad51-NTD to LacZ-NVH did not result in a significant increase in transcription.

The autonomous protein-expression-enhancing function of Rad51-NTD is unlikely to be controlled during transcription or simply arise from plasmid copy number differences.

The effects of WT and mutant Rad51-NTD on β-galactosidase activities (A), plasmid DNA copy numbers (B), relative steady-state levels of LacZ-NVH mRNA normalized to ACT1 (actin) mRNA (C), and relative ratios of LacZ-NVH mRNA versus plasmid DNA copy number (D). The wild-type yeast cells were transformed with indicated CEN-ARS plasmids, respectively, to express WT and mutant Rad51-NTD-LacZ-NVH fusion proteins or LacZ-NVH alone under the control of the native RAD51 gene promoter (PRAD51). The relative quantification (RQ = 2-ΔΔϹT) values were determined to reveal the plasmid DNA copy number and steady-state levels of LacZ-NVH mRNA by g-qPCR and RT-qPCR, respectively. LacZ and ACT1 were selected as target and reference protein-encoding genes, respectively, in both g-qPCR and RT-qPCR. The data shown represent mean ± SD from three independent biological data-points.

The protein quality control system moderately regulates autonomous PEE activities

The protein quality control system is a mechanism by which cells monitor proteins to ensure that they are appropriately folded (57). In the current study, we compared the protein steady-state levels (Figure 3A) and β-galactosidase activities (Figure 3B-3D) of Rad51-NTD-LacZ-NVH and LacZ-NVH in WT, hsp104Δ, new1Δ, doa1Δ, doa4Δ, san1Δ and oaz1Δ yeast cell lines. The protein products encoded by each of the six genes deleted from the latter mutant lines are all functionally relevant to protein homeostasis or prion propagation. Hsp104 is a heat-shock protein with disaggregase activities that disrupts protein aggregation (58, 59). New1 is a translation factor that fine-tunes ribosome recycling and the efficiency of translation termination (54). Doa1 (also called Ufd3) is an ubiquitin- and Cdc48-binding protein with a role in ubiquitin homeostasis and/or protein degradation (60, 61). The doa1Δ mutant exhibits diminished formation of [PSI+] prion (62). Doa4 is a deubiquitinating enzyme required for recycling ubiquitin from proteasome-bound ubiquitinated intermediates (63). The doa4Δ mutant exhibits increased sensitivity to the protein synthesis inhibitor cycloheximide (64). San1 is an ubiquitin-protein ligase that targets highly aggregation-prone proteins (65, 66). Oaz1 (ornithine decarboxylase antizyme) stimulates ubiquitin-independent degradation of Spe1 ornithine decarboxylase by the proteasome (67). We found that the β-galactosidase activities of Rad51-NTD-LacZ-NVH in WT and all six of the gene-knockout strains we examined were 10- to 29-fold higher than those of LacZ-NVH (Figure 3B). Intriguingly, the β-galactosidase activities of LacZ-NVH in the six gene-knockout mutants are all lower (30-70%) than those in WT (Figure 3C). In contrast, the β-galactosidase activities of Rad51-NTD-LacZ-NVH in WT are either slightly higher or lower than those in the six null mutants (Figure 3D). These results indicate that the addition of Rad51-NTD to LacZ-NVH can abrogate the protein homeostasis defects caused by the loss of each of these six genes. For example, Rad51-NTD might compensate for the ribosome assembly and translation defects in new1Δ (54), as well as the cycloheximide-hypersensitive phenotype displayed by doa4Δ (64). Accordingly, the β-galactosidase activities of Rad51-NTD-LacZ-NVH in the new1Δ and doa4Δ lines are higher than those in the WT, respectively. In contrast, the β-galactosidase activities of LacZ-NVH in the new1Δ and doa4Δ lines are lower, respectively, than those of WT. Finally, although the doa1Δ mutant is defective in [PSI+] prion formation (62), the steady-state levels of Rad51-NTD-LacZ-NVH in the doa1Δ line are also slightly higher than those in WT.

The expression-promoting function of Rad51-NTD is controlled during protein translation and does not affect ubiquitin-mediated protein degradation.

(A) The steady-state protein levels of Rad51-NTD-LacZ-NVH and LacZ-NVH in WT and six protein homeostasis gene knockout mutants. (B-D) The impact of six protein homeostasis genes on the β-galactosidase activity ratios of Rad51-NTD-LacZ-NVH to LacZ-NVH in WT and the six gene knockout mutants (B). The β-galactosidase activities of LacZ-NVH (C) and Rad51-NTD-LacZ-NVH (D) in WT and the six gene knockout mutants are shown. Asterisks indicate significant differences, with P values calculated using a two-tailed t-test (***, P value <0.001; **, P value <0.01; *, P value <0.05).

The N-end rule is not likely relevant to the PEE function of Q-rich motifs

The N-end rule links the in vivo half-life of a protein to the identity of its N-terminal residues. In S. cerevisiae, the N-end rule operates as part of the ubiquitin system and comprises two pathways. First, the Arg/N-end rule pathway, involving a single N-terminal amidohydrolase Nta1, mediates deamidation of N-terminal asparagine (N) and glutamine (Q) into aspartate (D) and glutamate (E), which in turn are arginylated by a single Ate1 R-transferase, generating the Arg/N degron. N-terminal R and other primary degrons are recognized by a single N-recognin Ubr1 in concert with ubiquitin-conjugating Ubc2/Rad6. Ubr1 can also recognize several other N-terminal residues, including lysine (K), histidine (H), phenylalanine (F), tryptophan (W), leucine (L) and isoleucine (I) (68–70). Second, the Ac/N-end rule pathway targets proteins containing N-terminally acetylated (Ac) residues. Prior to acetylation, the first amino acid methionine (M) is catalytically removed by Met-aminopeptidases (MetAPs), unless a residue at position 2 is non-permissive (too large) for MetAPs. If a retained N-terminal M or otherwise a valine (V), cysteine (C), alanine (A), serine (S) or threonine (T) residue is followed by residues that allow N-terminal acetylation, the proteins containing these AcN degrons are targeted for ubiquitylation and proteasome-mediated degradation by the Doa10 E3 ligase (71).

For two reasons, the PEE activities of these Q-rich domains are unlikely to arise from counteracting the N-end rule. First, the first two amino acid residues of Rad51-NTD, Hop1-SCD, Rad53-SCD1, Sup35-PND, Rad51-ΔN, and LacZ-NVH are MS, ME, ME, MS, ME, and MI, respectively, where M is methionine, S is serine, E is glutamic acid and I is isoleucine. Second, Sml1-NTD behaves similarly to these N-terminal fusion tags, despite its methionine and glutamine (MQ) amino acid signature at the N-terminus.

The relationship between PEE function, amino acid contents and structural flexibility

We applied an alanine scanning mutagenesis approach to reduce the percentages of S, T, Q or N in Rad51-NTD, Rad53-SCD1, and Sup35-NPD, respectively. These three Q-rich motifs exhibit a very strong positive relationship between STQ and STQN amino acid percentages and β-galactosidase activities (Figure 4 and Figure 5). IUPred2A (https://iupred2a.elte.hu/plot_new), a web-server for identifying disordered protein regions (72), also revealed that Rad51-NTD, Rad53-SCD1 and Sup35-NPD are structurally flexible peptides. These results are consistent with the notion that, due to high STQ or STQN content, SCDs or Q-rich motifs are intrinsically disordered regions (IDRs) in their native states, rather than adopting stable secondary and/or tertiary structures (4), and that a common feature of IDRs is their high content of S, T, Q, N, proline (P), glycine (G) and charged amino acids (17, 19, 20).

Relative β-galactosidase (LacZ) activities are correlated with the percentage STQ or STQN amino acid content of three Q-rich motifs.

(A) List of N-terminal tags with their respective length, numbers of S/T/Q/N amino acids, overall STQ or STQN percentages, and relative β-galactosidase activities. (B-D) Linear regressions between relative β-galactosidase activities and overall STQ or STQN percentages for Rad51-NTD (B), Rad53-SCD1 (C) and Sup35-PND (D). The coefficients of determination (R2) are indicated for each simple linear regression. (E) The amino acid sequences of wild-type and mutant Rad51-NTD, Rad51-SCD1 and Sup35-PND, respectively.

Alanine scanning mutagenesis of intrinsically disordered regions (IDRs).

The amino acid sequences of WT and mutant IDRs are listed in Table S1. Total protein lysates prepared from yeast cells expressing Rad51-NTD-LacZ-NVH (A), Sup35-PND-LacZ-NVH (B) or Rad53-SCD1-LacZ-NVH (C) were visualized by immunoblotting with anti-V5 antisera. Hsp104 was used as a loading control. Quantitative yeast β-galactosidase (LacZ) assays were carried out as described in Figure 1. Error bars indicate standard deviation between experiments (n = 3). Asterisks indicate significant differences when compared to LacZ-NVH, with p values calculated using a two-tailed t-test (**, P value <0.01 and ***, P value <0.001).

It is important to note that the threshold of STQ or STQN content varies in the three cases presented herein (Figure 4B). Thus, the percentage of STQ or STQN residues is not likely the only factor contributing to protein expression levels. Since G, P and glutamate (E) are enriched by >10% in Rad51-NTD, Rad53-SCD1 and Sup35-NPD, these three amino acids may also contribute to the PEE activities and structural flexibility of these three Q-rich motifs. Given that IDRs can endow proteins with structural and functional plasticity (73, 74), we hypothesized that Q-rich motifs (e.g., SCD, polyQ and polyQ/N) represent useful toolkits for creating new diversity during protein evolution.

Comparative proteome-wide analyses of amino acid contents, SCDs and polyX motifs

Next, we designed four JavaScript software programs (AS-aa-content, AS-codon-usage, AS-Finder-SCD and AS-Finder-polyX) for proteome-wide analyses (Table S4). AS-aa-content and AS-codon-usage determine the proteome-wide average contents of 20 different amino acids and the proteome-wide usage frequency of 64 genetic codons, respectively. ASFinder-SCD and ASFinder-polyX were applied to search for amino acid sequences that contain ≥3 S/T-Q motifs within a stretch of ≤100 residues (5) and for the polyX motifs of 20 different amino acids, respectively. In the latter case, diverse thresholds have been used in different studies or databases to define and detect polyX motifs (39, 75–77). Based on a previous study (39), we applied seven different thresholds to seek both short and long, as well as pure and impure, polyX strings in 20 different representative near-complete proteomes, including 4X (4/4), 5X (4/5-5/5), 6X (4/6-6/6), 7X (4/7-7/7), 8-10X (≥50%X), 11-10X (≥50%X) and ≥21X (≥50%X). The lowest threshold was ≥4/7, i.e., a minimum number of four identical X amino acid residues in a localized region of seven amino acid residues (Figures 6, Figures S1-S3 and Datasets DS7-DS32).

Percentages of proteins with different numbers of SCDs, and polyQ, polyQ/N or polyN tracts in 37 different eukaryotes.

We then searched and compared the near-complete proteomes of 26 different eukaryotes (Table 1), including the budding yeast S. cerevisiae, three pathogenic species of Candida, three filamentous ascomycete fungi (Neurospora crassa, Magnaporthe oryzae and Trichoderma reesei), three basidiomycete fungi (Cryptococcus neoformans, Ustilago maydis and Taiwanofungus camphoratus), the slime mold Dictyostelium discoideum, the malaria-causing unicellular protozoan parasite Plasmodium falciparum, six unicellular ciliates (Tetrahymena thermophila, Paramecium tetraurelia, Oxytricha trifallax, Stylonychia lemnae, Pseudocohnilembus persalinus and Stentor coeruleus), the fly Drosophila melanogaster, the mosquito Aedes aegypti, the nematode Caenorhabditis elegans, the zebrafish Danio rerio, the mouse Mus musculus, Homo sapiens, the higher plant Arabidopsis thaliana, and the single-celled green alga Chlamydomonas reinhardtii. The Benchmarking Universal Single-Copy Ortholog (BUSCO) scores of the near-universal single-copy gene orthologs of all 27 proteomes are 92.4-100% (Table 1). Genome or protein matrix scores >95% for model organisms are generally deemed complete reference genomes or proteomes (78).

It was reported previously that SCDs are overrepresented in the yeast and human proteomes (5, 79), and that polyX prevalence differs among species (43, 80–82). Our results reveal that the percentages of SCD proteins in the near-complete proteomes of 21 non-ciliate species and 6 ciliates range from 8.0% in P. falciparum, 13.9% in H. sapiens, 16.8% in S. cerevisiae, 24.2% in U. maydis, to a maximum of 58.0% in O. trifallax (Figure 6 and DS3). Among the 6050 proteins in the most recently updated S. cerevisiae reference proteome (https://www.uniprot.org/proteomes/UP000002311), we identified 1016 SCD-hosting proteins (DS3), including all 436 SCD-harboring proteins previously revealed by ScanProsite (5). ScanProsite is a publicly available database of protein families, domains and motifs (83).

The most striking finding in our study is that, due to their usage of the two noncanonical codons (UAAQ and UAGQ), Q (but not S, T or N) is used more frequently in five unicellular ciliates (i.e., T. thermophila, P. tetraurelia, O. trifallax, S. lemnae and P. persalinus) than in eukaryotes with standard genetic codons, including the unicellular ciliate S. coeruleus and all of the 20 non-ciliate species we examined herein (Figures S1 and DS2). Hereafter, we refer to the five unicellular ciliates with reassigned stop codons as ‘group I’ ciliates. Due to higher proteome-wide Q contents, there are higher percentages of SCD, polyQ and polyQ/N in the five group I ciliates than in S. coeruleus (Figures 6, S2, S3, and DS2-DS4).

Next, we analyzed the SCD and polyX proteins encoded by the transcriptomes of 11 different ciliate species. These transcripts were originally generated as part of the Marine Microbial Eukaryote Transcriptome Sequencing Project (MMETSP) (84), which were then reassembled and reannotated by Brown and colleagues (85). All transcripts are publicly available from Zendo (https://zenodo.org/record/1212585#.Y79zoi2l3PA) (85). We applied TransDecoder (https://github.com/TransDecoder/TransDecoder/wiki) to identify candidate coding regions within the transcript sequences. Five of those eleven ciliates have reassigned UAAQ and UAGQ codons (hereafter termed “group II ciliates”), i.e., Aristerostoma spp., Favella ehrenbergii, Pseudokeronopsis spp., Strombidium inclinatum, and Uronema spp.. Like sessile S. coeruleus, group III ciliates (Climacostomum virens, Litonotus pictus and Protocruzia adherens) possess the standard genetic codes. Group IV ciliates encompass Mesodinium pulex, Blepharisma japonicum and Euplotes focardii, each of which utilizes the reassigned codons UAAY, UAGY, UGAW and UGAC, respectively (Table 1). For two reasons, the proteins encoded by these MMESTP transcripts are unlikely to represent the entire protein complement of all 11 ciliate species. First, many MMETSP transcripts are not intact (i.e., broken mRNAs) and thus encode incomplete protein sequences. Second, except for C. virens (94.7%), the BUSCO protein scores of these MMESTP transcripts only range from 52.6% to 89.9% (Table 1). Nevertheless, our results indicate that Q is used more frequently in group I and group II ciliates than in group III and group IV ciliates or in a further 20 non-ciliate species (Figure S1 and DS2). Accordingly, proportions of SCD, polyQ and polyQ/N proteins in all group I and group II ciliates are higher than they are in the three group III ciliates (except L. pictus) and the three group IV ciliates, respectively. Since N is used more frequently in L. pictus than the other ciliates in groups II-IV, it has higher percentages of polyN and polyQ/N proteins (Figures 6, S2, S3, and DS2-DS4). Our data also indicates that Y, W or C are not used more frequently in the three group IV ciliates than in the other 14 ciliate or 20 non-ciliate species (Figure S1 and DS2). Reassignments of stop codons to Y, W or C also do not result in higher percentages of polyY, polyW or polyC proteins in the three group IV ciliates, respectively (Figure S3).

To further confirm the above-described results, we normalized the runs of amino acids and created a null expectation from each proteome by determining the ratios of the overall number of X residues for each of the seven polyX motifs relative to those in the entire proteome of each species, respectively. The results for four different polyX motifs, i.e., polyQ, polyN, polyS and polyT, are presented in Figures 7-10 and DS5. The results summarized in Figures 7-10 support that polyX prevalence differs among species and that the overall X contents of polyX motifs often but not always correlate with the X usage frequencies in entire proteomes (43). Most importantly, our results reveal that, compared to S. coeruleus or several non-ciliate eukaryotic organisms (e.g., P. falciparum, C. elegans, D. rerio, M. musculus and H. sapiens), the five ciliates with reassigned TAAQ and TAGQ codons not only have higher Q usage frequency but also more polyQ motifs in their proteomes (Figure 7 and DS5). In contrast, polyQ motifs prevail in C. albicans, C. tropicalis, D. discoideum, C. reinhardtii, D. melanogaster and A. aegypti, though the Q usage frequencies in their entire proteomes are not significantly higher than those of other eukaryotes (Figure 7 and DS5). Due to their higher N usage frequencies, D. discoideum, P. falciparum and P. persalinus have more polyN motifs than the other 23 eukaryotes we examined here (Figure 8 and DS5). Generally speaking, all 26 eukaryotes we assessed have similar S usage frequencies and percentages of S contents in polyS motifs (Figure 9 and DS5). Among these 26 eukaryotes, D. discoideum possesses many more polyT motifs, though its T usage frequency is similar to that of the other 25 eukaryotes (Figure 10 and DS5). Several other polyX motifs are particularly enriched in specific eukaryotes, for example, polyK and polyY in P. falciparum, polyK and polyF in D. discoideum, polyG, polyA, polyP and polyW in C. reinhardtii, as well as the longest polyC (i.e., ≥21C and ≥50%C) in C. tropicalis (DS5). Further investigations will decipher the structural and functional relevance of those polyX motif proteins. In conclusion, these normalized results further confirm that reassignment of stop codons to Q indeed results in both higher Q usage frequencies and more polyQ motifs in ciliates.

Q contents in 7 different types of polyQ motifs in 20 near-complete proteomes.

The five ciliates with reassigned stops codon (TAAQ and TAGQ) are indicated in red. Stentor coeruleus, a ciliate with standard stop codons, is indicated in green.

N contents in 7 different types of polyN motifs in 20 near-complete proteomes.

The five ciliates with reassigned stops codon (TAAQ and TAGQ) are indicated in red. Stentor coeruleus, a ciliate with standard stop codons, is indicated in green.

S contents in 7 different types of polyS motifs in 20 near-complete proteomes.

The five ciliates with reassigned stops codon (TAAQ and TAGQ) are indicated in red. Stentor coeruleus, a ciliate with standard stop codons, is indicated in green.

T contents in 7 different types of polyT motifs in 20 near-complete proteomes.

The five ciliates with reassigned stops codon (TAAQ and TAGQ) are indicated in red. Stentor coeruleus, a ciliate with standard stop codons, is indicated in green.

The frequency of TAAQ and TAGQ, CAAQ and CAGQ usage in 26 different organisms

PolyQ motifs have a particular length-dependent codon usage that relates to strand slippage in CAG/CTG trinucleotide repeat regions during DNA replication (37, 39). In most organisms having standard genetic codons, Q is encoded by CAGQ and CAAQ. We applied AS-Xcontent-7polyX, a JavaScript software program (Table S4), to determine and compare proteome-wide Q contents, as well as CAGQ usage frequencies (i.e., the ratio between CAGQ and the sum of CAGQ, CAGQ, TAAQ, and TAGQ) (Table 2 and DS6). Our results reveal that the likelihood of forming long CAG/CTG trinucleotide repeats is higher in five eukaryotes due to their higher CAGQ usage frequencies, including D. melanogaster (86.6% Q), D. rerio (74.0% Q), M. musculus (74.0% Q), H. sapiens (73.5% Q), and C. reinhardtii (87.3% Q) (orange background, Table 2). In contrast, another five eukaryotes that possess high numbers of polyQ motifs (i.e., D. discoideum, C. albicans, C. tropicalis, P. falciparum and S. coeruleus) (Figure 7) utilize more CAAQ (96.2%, 84.6%, 84.5%, 86.7% and 75.7%) than CAGQ (3.8%, 15.4%, 15.5%, 13.3% and 24.3%), respectively, to avoid forming long CAG/CTG trinucleotide repeats (green background, Table 2). Similarly, all five ciliates with reassigned stop codons (TAAQ and TAGQ) display low CAGQ usage frequencies (i.e., ranging from 3.8% Q in P. persalinus to 12.6% Q in O. trifallax) (Table 2). Accordingly, the CAG-slippage mechanism may operate more frequently in C. reinhardtii, D. melanogaster, D. rerio, M. musculus and H. sapiens than in D. discoideum, C. albicans, C. tropicalis, P. falciparum, S. coeruleus and the five ciliates with reassigned stop codons (TAAQ and TAGQ).

Q-rich-motif proteins are overrepresented in specialized biological processes of various eukaryotic proteomes

To determine the biological impacts of Q-rich-motif proteins, we designed a JavaScript software tool AS-GOfuncR-FWER (Table S4) to carry out comparative Gene Ontology (GO) enrichment analyses using information on the functions of genes provided by the GO knowledgebase (http://geneontology.org). Rigorous statistical testing for overrepresentation or underrepresentation of SCD and polyX proteins was performed using GOfuncR (https://bioconductor.org/packages/release/bioc/html/GOfuncR.html), an R package program that conducts standard candidate vs. background enrichment analysis employing the hypergeometric test. The raw p-values were adjusted according to the Family-Wise Error Rate (FWER). The same method was applied to the GO enrichment analysis of human genomes (86). The results presented in Figure 11 and Figure 12 (DS7-DS32) support the hypothesis that Q-rich motifs prevail in proteins involved in specialized biological processes, including S. cerevisiae RNA-mediated transposition, C. albicans filamentous growth, peptidyl-glutamic acid modification in ciliates with reassigned stop codons (TAAQ and TAGQ), T. thermophila xylan catabolism, D. discoideum sexual reproduction, P. falciparum infection, as well as the nervous systems of D. melanogaster, M. musculus, and H. sapiens. In contrast, peptidyl-glutamic acid modification is not overrepresented with Q-rich-motif proteins in S. coeruleus, a ciliate with standard stop codons.

Selection of biological processes with overrepresented SCD-containing proteins in different eukaryotes.

The percentages and number of SCD-containing proteins in our search that belong to each indicated Gene Ontology (GO) group are shown. GOfuncR (86) was applied for GO enrichment and statistical analysis. The p values adjusted according to the Family-wise error rate (FWER) are shown.

Selection of biological processes with overrepresented polyQ-containing proteins in different eukaryotes.

The percentages and numbers of polyQ-containing proteins in our search that belong to each indicated Gene Ontology (GO) group are shown. GOfuncR (86) was applied for GO enrichment and statistical analysis. The p values adjusted according to the Family-wise error rate (FWER) are shown. The five ciliates with reassigned stops codons (TAAQ and TAGQ) are indicated in red. Stentor coeruleus, a ciliate with standard stop codons, is indicated in green.

Our results are also consistent with a previous report that there is an overrepresentation of conserved Q-rich-motif proteins in pathways related to the human nervous system (79). For instance, human CTTNBP2 (1663 amino acid residues) is a neuron-specific F-actin-associated SCD protein that is involved in the formation and maintenance of dendritic spines and it is associated with autism spectrum disorders (87, 88). Human CTTNBP2 possesses ten S/T-Q motifs (T466Q, T493Q, S1580Q, S553Q, S634Q, S994Q, S1392Q, S1580Q, T1621Q and S1624Q). Mouse CTTNBP2 has 630 amino acid residues, four S/T-Q motifs (S419Q, T463Q, S550Q, S624Q), and it shares a high amino acid identity with the N-terminus (1-640 amino acid residues) of human CTTNBP2. IUPred2A (https://iupred2a.elte.hu/plot_new) also reveals that both human CTTNBP2 (220-1633 residues) and mouse CTTNBP2 (220-630 residues) are Q/N-rich IDRs with high percentages of S, T, Q, N, G, P, R and K. We reported recently that mouse CTTNBP2 forms self-assembled condensates through its C-terminal IDR and it facilitates co-condensation of an abundant excitatory postsynaptic scaffold protein SHANK3 at dendritic spines in a Zn2+-dependent manner (89).

Q-rich-motif proteins prevail in the T. thermophila xylan catabolic process

The proteome of T. thermophila contains 58 proteins involved in the xylan catabolic process (GO ID: 45493), of which 56 (97%), 55 (95%), 58 (100%), and 49 (84%) proteins harbor SCD, polyQ, polyQ/N, and polyN tracts, respectively (DS19-DS24). Using the NCBI BLASTP search tool with an expect value (E-value) ≤10e-5 to search for homologs of these 58 proteins among all the other 16 ciliates analyzed in this study, we only identified 144 proteins with amino acid identity >60% and a raw alignment score of >150 (Dataset DS33). Thus, T. thermophila has more abundant xylan catabolic proteins than all other ciliates we examined herein.

Most proteins involved in T. thermophila meiosis harbor one or more Q-rich motif(s)

Ciliate meiosis is remarkable relative to that of other studied sexual eukaryotes. Ciliates often have two types of nuclei. Their diploid micronucleus (MIC) carries the cell germline, the genetic material of which is inherited via sexual reproduction and meiosis. The polyploid macronucleus (MAC) or vegetative nucleus provides nuclear RNA for vegetative growth. The MAC is generated from the MIC by massive amplification, editing and rearrangement of the genome (see reviews in (90, 91)). In T. thermophila, the most intensively studied ciliate, meiotic MICs undergo extreme elongation (by ∼50-fold) and form proteinaceous condensates called “crescents”. Within these elongated crescents, telomeres and centromeres of all meiotic chromosomes are rearranged at opposing ends in a stretched bouquet-like manner. Meiotic pairing and recombination take place within the crescents (see review in (92)). It has been reported that ATR1 (Ataxia Telangiectasia Mutated 1), an evolutionarily conserved DNA damage senor protein kinase, senses Spo11-induced DSBs and triggers the elongation of MICs (93). Meiosis-specific CYC2 and CYC17 cyclins, as well as cyclin-dependent kinase CDK3, are required to initiate meiosis and for crescent assembly (94–96). CYC2/CDK2 promotes bouquet formation in MICs by controlling microtubule-directed elongation (96) and it also controls the gene expression of proteins involved in DSB formation (SPO11), DNA repair (COM1, EXO1, DMC1), and crossover formation (HOP2, MND1, MSH4, MSH5, ZPH3, BIM1, and BIM2) (97). The DPL2/E2fl1 complex, a meiosis-specific transcription factor, promotes transcriptional induction of DNA repair proteins and chromosome-associated structural proteins, including MRE11, COM1, EXO1, RAD50, RAD51, SMC1, SMC2, SMC3, SMC4, REC8, ESP1, and CNA1, among others (97). Nevertheless, the molecular mechanisms underlying crescent assembly and disassembly remain poorly understood.

Among the 124 T. thermophila meiotic proteins (DS34, see review in (92)), we identified 85 SCD proteins, 54 polyQ proteins, 106 polyQ/N proteins, 32 polyN proteins, 32 polyS proteins, and 48 polyK proteins, respectively. Notably, there are 48 and 59 meiotic proteins that contain ≥4 SCDs and/or ≥4 polyQ/N tracts, respectively. For instance, DPL2, CYC2, CYC17, and ATR1 each contain 15, 6, 8 and 5 SCDs, 2, 0, 1 and 2 polyQ tracts, as well as 11, 4, 4 and 5 polyQ/N tracts, respectively. Pars11, a chromatin-associated protein required for inducing and limiting Spo11-induced DSBs, has 14 SCDs, 2 polyQ tracts and 6 polyQ/N tracts. Spo11-induced DSBs promote ATR1-dependent Pars11 phosphorylation and its removal from chromatin (97). Many T. thermophila meiotic DSB repair proteins also harbor several SCDs, polyQ tracts and/or polyQ/N tracts, including MSH4, MSH5, SGS1, FANCM, REC8, and ZPH3, among others.

Several T. thermophila proteins involved in editing and rearrangement of the MIC genome also harbor multiple Q-rich motifs. PDD1 (programmed DNA degradation 1), a conjugation-specific HP1-like protein, has 13 SCDs, 3 polyQ tracts and 6 polyQ/N tracts. Mutations in the chromodomain or the chromoshadow domain of PDD1 were found previously to elicit PDD1 mislocalization, prevented histone H3 dimethylation on K9, abolished removal of internal eliminated sequences (IES), and/or resulted in the production of inviable progeny (98). DCL1 (Dicer-like 1) has 5 SCDs, 1 polyQ tract and 8 polyQ/N tracts. DCL1 is required for processing the MIC transcripts to siRNA-like scan (scn) RNAs, as well as for methylation of histone H3 at Lys 9. This latter modification occurs specifically on sequences (IESs) to be eliminated (99). GIW1 (gentlemen-in-waiting 1) physically directs a mature Argonaute-siRNA complex to the MIC nucleus, thus promoting programmed IES elimination (100). GIW1 has 9 SCDs and 2 polyQ/N tracts (DS34).

Using IUPred2A (https://iupred2a.elte.hu/plot_new), we found that the Q-rich motifs in most (if not all) of these meiotic proteins are intrinsically disordered. Accordingly, we speculate that, like the C-terminal IDR of mammalian CTTNBP2, the Q-rich motifs in T. thermophila meiotic proteins might form tunable proteinaceous condensates to regulate assembly and disassembly of the “crescents” in ciliate MICs.

Discussion

We present three unexpected results in this report. First, the Q-rich motifs of several yeast proteins (Rad51-NTD, Rad53-SCD1, Hop1-SCD, Sml1-NTD, Sup35-PND, Ure2-UPD, and New1-NPD) all exhibit autonomous PEE activities. These structurally flexible Q-rich motifs have useful potential for applications in both basic research (e.g., synthetic biology) and biotechnology. Further investigations would prove illuminating as to how these Q-rich motifs exert this PEE function in yeast and whether Q-rich motifs in other eukaryotes also possess similar PEE activities. Second, the reassignment of stop codons to Q in the group I and group II ciliates significantly increases proteome-wide Q usage, leading to massive expansion of structurally flexible or even intrinsically disordered Q-rich motifs. In contrast, reassignments of stop codons to Y, W or C do not result in higher usages of these three amino acid residues, nor higher percentages of W-, Y- or C-rich proteins in the three group IV ciliates, respectively. These results are consistent with the notion that, unlike for Q, the Y, W and C residues are not common in IDRs (17, 19, 20). Third, the results in Table 2 support that a decrease or increase of CAGQ usage frequency in different eukaryotes is responsible for a reduction or augmentation, respectively, of polyQ instability (or expansion) caused by DNA strand slippage in CAG/CTG trinucleotide repeat regions during DNA replication (37, 39). Accordingly, it would be interesting to decipher how the molecular mechanism(s) that controls the codon usage bias of TAAQ, TAGQ, CAAQ and CAGQ evolved.

Due to their structural flexibility, Q-rich motifs can endow proteins with structural and functional plasticity. Based on previous reports (73, 74) and our findings from this study, Q-rich motifs, such as IDRs, are useful toolkits for generating novel diversity during protein evolution, including by enabling greater protein expression, protein-protein interactions, posttranslational modifications, increased solubility, and tunable stability, among other important traits. This speculation may explain three intriguing phenomena. First, due to higher Q usage, many proteins involved in evolutionarily conserved biological processes in group I and group II ciliates display more diverse amino acid sequences than the respective proteins in other ciliate or non-ciliate species. Accordingly, it is sometimes difficult to identify authentic protein homologs among different ciliates, particularly for group I and group II ciliates. We highlight the example of the 58 proteins involved in xylan catabolysis in T. thermophila (Dataset DS33). Second, our GO enrichment results reveal that Q-rich motifs prevail in proteins involved in specialized biological processes (Figure 11 and Figure 12). In theory, structurally flexible Q-rich motifs might form various membraneless organelles or proteinaceous condensates via intracellular liquid-liquid phase separation in tunable manners, including protein posttranslational modification, protein-protein interaction, protein-ligand binding, among other processes (21, 22). A typical example is that the C-terminus of mouse CTTNBP2 facilitates co-condensation of CTTNBP2 with SHANK3 at the postsynaptic density in a Zn2+-dependent manner (89). Third, Borgs are long and linear extrachromosomal DNA sequences in methane-oxidizing Methanoperedens archaea, which display the potential to augment methane oxidation (101). A striking feature of Borgs is pervasive tandem direct repeat (TR) regions. TRs in open reading frames (ORFs) are under very strong selective pressure, leading to perfect amino acid TRs (aaTRs) that are commonly IDRs. Notably, aaTRs often contain disorder-promoting amino acids, including Q, N, P, T, E, K, V, D and S (102). Accordingly, distinct evolutionary strategies are employed in different species to alter protein regions that are structurally flexible or even intrinsically disordered. Further investigations are needed to determine if liquid-liquid phase separation prevails in ciliates with reassigned TAAQ and TAGQ codons and/or in the specialized biological processes in various species we have described herein.

Conclusions

One of the most interesting questions in genome diversity is why many ciliates reassign their nuclear stop codons into amino acids, e.g., glutamine (Q), tyrosine (Y), tryptophan (W) or cysteine (C). The impacts of such genome-wide alternations had not been well understood. Here, we show that glutamine (Q) is used more frequently in all 10 ciliate species possessing reassigned TAAQ and TAGQ codons than in other ciliates and non-ciliate species. The consequence of this preponderance of Q is the massive expansion of proteins harboring structurally flexible or even intrinsically disordered Q-rich motifs. Since Q-rich motifs can endow proteins with structural and functional plasticity and Q-rich-motif proteins are overrepresented in several species-specific or even phylum-specific biological processes, we suggest that Q-rich motifs are useful toolkits for evolutionary novelty.

Supporting information